---
title: "Hands-on Exercise 9 (Part 1)"
author: "Bui Tien Thanh (tt.bui.2025@mitb.smu.edu.sg)"
date: "2026-06-16"
date-modified: "last-modified"
description: "Choropleth Mapping with R"
format: html
execute:
  echo: true
  eval: true
  warning: false
  freeze: true
---

# Learning Outcome

Choropleth mapping involves the symbolisation of enumeration units, such as countries, provinces, states, counties or census units, using area patterns or graduated colors. For example, a social scientist may need to use a choropleth map to portray the spatial distribution of aged population of Singapore by Master Plan 2014 Subzone Boundary.

In this chapter, you will learn how to plot functional and truthful choropleth maps by using an R package called **tmap**.

# Getting Started

In this hands-on exercise, the key R package used is the **tmap** package in R. Beside tmap, four other R packages will be used:

- **readr** for importing delimited text files,
- **tidyr** for tidying data,
- **dplyr** for wrangling data, and
- **sf** for handling geospatial data.

Among the four packages, readr, tidyr and dplyr are part of the **tidyverse** package.

The code chunk below will be used to install and load these packages in RStudio.

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

# Importing Data into R

## The Data

## Importing Geospatial Data into R

```{r}
mpsz <- st_read(dsn = "../data/geospatial/MP14_SUBZONE_WEB_PL",
                layer = "MP14_SUBZONE_WEB_PL")
```

You can examine the content of `mpsz` by using the code chunk below.

```{r}
mpsz
```

## Importing Attribute Data into R

Next, we will import respopagsex2011to2020.csv file into RStudio and save the file into an R dataframe called `popagsex`.

The task will be performed by using `read_csv()` function of readr package as shown in the code chunk below.

```{r}
popdata <- read_csv("../data/aspatial/respopagesextod2011to2020.csv")
```

## Data Preparation

Before a thematic map can be prepared, you are required to prepare a data table with year 2020 values. The data table should include the variables PA, SZ, YOUNG, ECONOMY ACTIVE, AGED, TOTAL, DEPENDENCY.

- **YOUNG**: age group 0 to 4 until age group 20 to 24,
- **ECONOMY ACTIVE**: age group 25–29 until age group 60–64,
- **AGED**: age group 65 and above,
- **TOTAL**: all age groups, and
- **DEPENDENCY**: the ratio between young and aged against the economy active group.

### Data wrangling

The following data wrangling and transformation functions will be used:

- `pivot_wider()` of tidyr package, and
- `mutate()`, `filter()`, `group_by()` and `select()` of dplyr package.

```{r}
popdata2020 <- popdata %>%
  filter(Time == 2020) %>%
  group_by(PA, SZ, AG) %>%
  summarise(`POP` = sum(`Pop`)) %>%
  ungroup() %>%
  pivot_wider(names_from=AG,
              values_from=POP) %>%
  mutate(YOUNG = rowSums(.[3:6])
         +rowSums(.[12])) %>%
mutate(`ECONOMY ACTIVE` = rowSums(.[7:11])+
rowSums(.[13:15]))%>%
mutate(`AGED`=rowSums(.[16:21])) %>%
mutate(`TOTAL`=rowSums(.[3:21])) %>%
mutate(`DEPENDENCY` = (`YOUNG` + `AGED`)
/`ECONOMY ACTIVE`) %>%
  select(`PA`, `SZ`, `YOUNG`,
       `ECONOMY ACTIVE`, `AGED`,
       `TOTAL`, `DEPENDENCY`)
```

### Joining the attribute data and geospatial data

Before we can perform the georelational join, one extra step is required to convert the values in PA and SZ fields to uppercase. This is because the values of PA and SZ fields are made up of upper- and lowercase. On the other hand, SUBZONE_N and PLN_AREA_N are in uppercase.

```{r}
popdata2020 <- popdata2020 %>%
  mutate(across(c(PA, SZ), toupper)) %>%
  filter(`ECONOMY ACTIVE` > 0)
```

Next, `left_join()` of dplyr is used to join the geographical data and attribute table using planning subzone name e.g. SUBZONE_N and SZ as the common identifier.

```{r}
mpsz_pop2020 <- left_join(mpsz, popdata2020,
                          by = c("SUBZONE_N" = "SZ"))
```

::: {.callout-note}
`left_join()` of dplyr package is used with `mpsz` simple feature data frame as the left data table to ensure that the output will be a simple features data frame.
:::

```{r}
write_rds(mpsz_pop2020, "../data/rds/mpszpop2020.rds")
```

# Choropleth Mapping Geospatial Data Using tmap

Two approaches can be used to prepare a thematic map using tmap:

- Plotting a thematic map quickly by using `qtm()`.
- Plotting a highly customisable thematic map by using tmap elements.

## Plotting a choropleth map quickly by using qtm()

The easiest and quickest way to draw a choropleth map using tmap is `qtm()`. It is concise and provides a good default visualisation in many cases.

The code chunk below will draw a cartographic standard choropleth map as shown below.

```{r}
tmap_mode("plot")
qtm(mpsz_pop2020,
    fill = "DEPENDENCY")
```

::: {.callout-note}
- `tmap_mode()` with `"plot"` option is used to produce a static map. For interactive mode, the `"view"` option should be used.
- The `fill` argument is used to map the attribute (i.e. DEPENDENCY).
:::

## Creating a choropleth map by using tmap's elements

Despite the usefulness of `qtm()` for drawing a choropleth map quickly and easily, its disadvantage is that it makes aesthetics of individual layers harder to control. To draw a high quality cartographic choropleth map, tmap's drawing elements should be used.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_title("Distribution of Dependency Ratio by planning subzone") +
  tm_layout(frame = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS",
             position = c("left", "bottom"))
```

In the following sub-section, we will share tmap functions that are used to plot these elements.

## Drawing a base map

The basic building block of tmap is `tm_shape()` followed by one or more layer elements such as `tm_fill()` and `tm_polygons()`.

In the code chunk below, `tm_shape()` is used to define the input data (i.e. `mpsz_pop2020`) and `tm_polygons()` is used to draw the planning subzone polygons.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons()
```

## Drawing a choropleth map using tm_polygons()

To draw a choropleth map showing the geographical distribution of a selected variable by planning subzone, we just need to assign the target variable such as DEPENDENCY to `tm_polygons()`.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY")
```

::: {.callout-note}
Things to note from `tm_polygons()`:

- The default interval binning used to draw the choropleth map is called `"pretty"`. A detailed discussion of the data classification methods supported by tmap will be provided in sub-section 4.3.
- The default colour scheme used is **YlOrRd** of ColorBrewer. You will learn more about the color scheme in sub-section 4.4.
- By default, missing values will be shaded in grey.
:::

## Drawing a choropleth map using tm_fill() and tm_borders()

Actually, `tm_polygons()` is a wrapper of `tm_fill()` and `tm_borders()`. `tm_fill()` shades the polygons by using the default colour scheme and `tm_borders()` adds the borders of the shapefile onto the choropleth map.

The code chunk below draws a choropleth map by using `tm_fill()` alone.

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY")
```

Notice that the planning subzones are shaded according to the respective dependency values.

To add the boundary of the planning subzones, `tm_borders()` will be used as shown in the code chunk below.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons(fill = "DEPENDENCY") +
  tm_borders(lwd = 0.01,
             fill_alpha = 0.1)
```

Notice that light-gray border lines have been added on the choropleth map.

The `alpha` argument is used to define transparency — a number between 0 (totally transparent) and 1 (not transparent). By default, the alpha value of the col is used (normally 1).

Beside the alpha argument, there are three other arguments for `tm_borders()`:

- `col` = border colour,
- `lwd` = border line width. The default is 1, and
- `lty` = border line type. The default is `"solid"`.

## Data classification methods of tmap

Most choropleth maps employ some methods of data classification. The point of classification is to take a large number of observations and group them into data ranges or classes.

tmap provides a total of ten data classification methods: **fixed**, **sd**, **equal**, **pretty** (default), **quantile**, **kmeans**, **hclust**, **bclust**, **fisher**, and **jenks**.

To define a data classification method, the `style` argument of `tm_fill()` or `tm_polygons()` will be used.

### Plotting choropleth maps with built-in classification methods

The code chunk below shows a quantile data classification that used 5 classes.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "jenks",
        n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

In the code chunk below, equal data classification method is used.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "equal",
        n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

Notice that the distribution of the quantile data classification method is more evenly distributed than the equal data classification method.

::: {.callout-warning}
**Maps Lie!** Different classification methods can produce very different visual patterns from the same data. Always choose the method that best represents the underlying distribution.
:::

::: {.callout-tip title="DIY"}
Using what you have learned, prepare choropleth maps by using different classification methods supported by tmap and compare their differences.
:::

::: {.callout-tip title="DIY"}
Prepare choropleth maps using the same classification method but with different numbers of classes (i.e. 2, 6, 10, 20). Compare the output maps — what observation can you draw?
:::

### Plotting choropleth map with custom break

For all the built-in styles, the category breaks are computed internally. In order to override these defaults, the breakpoints can be set explicitly by means of the `breaks` argument to `tm_fill()`. It is important to note that, in tmap the breaks include a minimum and maximum. As a result, in order to end up with n categories, n+1 elements must be specified in the breaks option (the values must be in increasing order).

Before we get started, it is always a good practice to get some descriptive statistics on the variable before setting the break points. The code chunk below will be used to compute and display the descriptive statistics of the DEPENDENCY field.

```{r}
summary(mpsz_pop2020$DEPENDENCY)
```

With reference to the results above, we set break points at 0.60, 0.70, 0.80, and 0.90. In addition, we also need to include a minimum and maximum, which we set at 0 and 1.00. Our breaks vector is thus `c(0, 0.60, 0.70, 0.80, 0.90, 1.00)`.

Now, we will plot the choropleth map by using the code chunk below.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
          breaks = c(0, 0.60, 0.70, 0.80, 0.90, 1.00)) +
  tm_borders(fill_alpha = 0.5)
```

## Colour Scheme

### Using ColourBrewer palette

To change the colour, we assign the preferred colour to the `values` argument of `tm_scale_intervals()` as shown in the code chunk below.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "quantile",
        n = 5,
        values = "brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```

Notice that the choropleth map is shaded in green.

To reverse the colour shading, add a `"-"` prefix.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "quantile",
        n = 5,
        values = "-brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```

## Map Layouts

Map layout refers to the combination of all map elements into a cohesive map. Map elements include among others the objects to be mapped, the title, the scale bar, the compass, margins and aspect ratios. Colour settings and data classification methods covered in the previous section relate to the palette and break-points used to affect how the map looks.

### Map Legend

In tmap, several `tm_legend()` options are provided to change the placement, format and appearance of the legend.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "jenks",
        n = 5,
        values = "brewer.greens"),
      fill.legend = tm_legend(
        title = "Dependency ratio")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Distribution of Dependency Ratio by planning subzone \n(Jenks classification)")
```

### Map style

tmap allows a wide variety of layout settings to be changed. They can be called by using `tmap_style()`.

The code chunk below shows the classic style is used.

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "-Greens") +
  tm_borders(alpha = 0.5) +
  tmap_style("classic")
```

### Cartographic Furniture

Beside map style, tmap also provides arguments to draw other map furniture such as compass, scale bar and grid lines.

In the code chunk below, `tm_compass()`, `tm_scale_bar()` and `tm_grid()` are used to add compass, scale bar and grid lines onto the choropleth map.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_title("Distribution of Dependency Ratio by planning subzone") +
  tm_layout(frame = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS",
             position = c("left", "bottom"))
```

To reset the default style, refer to the code chunk below.

```{r}
tmap_style("white")
```

## Drawing Small Multiple Choropleth Maps

Small multiple maps, also referred to as facet maps, are composed of many maps arranged side-by-side, and sometimes stacked vertically. Small multiple maps enable the visualisation of how spatial relationships change with respect to another variable, such as time.

In tmap, small multiple maps can be plotted in three ways:

- by assigning multiple values to at least one of the aesthetic arguments,
- by defining a group-by variable in `tm_facets()`, and
- by creating multiple stand-alone maps with `tmap_arrange()`.

### By assigning multiple values to at least one of the aesthetic arguments

In this example, small multiple choropleth maps are created by defining ncols in `tm_fill()`.

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill(c("YOUNG", "AGED"),
          style = "equal",
          palette = "Blues") +
  tm_layout(legend.position = c("right", "bottom")) +
  tm_borders(alpha = 0.5) +
  tmap_style("white")
```

In this example, small multiple choropleth maps are created by assigning multiple values to at least one of the aesthetic arguments.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons(c("DEPENDENCY","AGED"),
          style = c("equal", "quantile"),
          palette = list("Blues","Greens")) +
  tm_layout(legend.position = c("right", "bottom"))
```

### By defining a group-by variable in tm_facets()

In this example, multiple small choropleth maps are created by using `tm_facets()`.

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "Blues",
          thres.poly = 0) +
  tm_facets(by="REGION_N",
            free.coords=TRUE) +
  tm_layout(legend.show = FALSE,
            title.position = c("center", "center"),
            title.size = 20) +
  tm_borders(alpha = 0.5)
```

### By creating multiple stand-alone maps with tmap_arrange()

In this example, multiple small choropleth maps are created by creating multiple stand-alone maps with `tmap_arrange()`.

```{r}
youngmap <- tm_shape(mpsz_pop2020)+
  tm_polygons("YOUNG",
              style = "quantile",
              palette = "Blues")

agedmap <- tm_shape(mpsz_pop2020)+
  tm_polygons("AGED",
              style = "quantile",
              palette = "Blues")

tmap_arrange(youngmap, agedmap, asp=1, ncol=2)
```

## Mapping Spatial Objects Meeting a Selection Criterion

Instead of creating small multiple choropleth maps, you can also use a selection function to map spatial objects meeting the selection criterion.

```{r}
tm_shape(mpsz_pop2020[mpsz_pop2020$REGION_N=="CENTRAL REGION", ])+
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "Blues",
          legend.hist = TRUE,
          legend.is.portrait = TRUE,
          legend.hist.z = 0.1) +
  tm_layout(legend.outside = TRUE,
            legend.height = 0.45,
            legend.width = 5.0,
            legend.position = c("right", "bottom"),
            frame = FALSE) +
  tm_borders(alpha = 0.5)
```

# Reference
