---
title: "Hands-on Exercise 9 (Part 2)"
author: "Bui Tien Thanh (tt.bui.2025@mitb.smu.edu.sg)"
date: "2026-06-16"
date-modified: "last-modified"
description: "Visualising Geospatial Point Data"
format: html
execute:
  echo: true
  eval: true
  warning: false
  freeze: true
---

# Learning Outcome

Proportional symbol maps (also known as graduate symbol maps) are a class of maps that use the visual variable of size to represent differences in the magnitude of a discrete, abruptly changing phenomenon, e.g. counts of people. Like choropleth maps, you can create classed or unclassed versions of these maps. The classed ones are known as range-graded or graduated symbols, and the unclassed are called proportional symbols, where the area of the symbols are proportional to the values of the attribute being mapped. In this hands-on exercise, you will learn how to create a proportional symbol map showing the number of wins by Singapore Pools' outlets using an R package called **tmap**.

By the end of this hands-on exercise, you will acquire the following skills by using appropriate R packages:

- Import an aspatial data file into R.
- Convert it into a simple point feature data frame and at the same time assign an appropriate projection reference to the newly created simple point feature data frame.
- Plot interactive proportional symbol maps.

# Getting Started

Before we get started, we need to ensure that the **tmap** package of R and other related R packages have been installed and loaded into R.

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

# Importing Data into R

## The Data

## Data Import and Preparation

```{r}
sgpools <- read_csv("../data/aspatial/SGPools_svy21.csv")
```

After importing the data file into R, it is important for us to examine if the data file has been imported correctly.

The code chunk below shows `list()` is used to do the job.

```{r}
list(sgpools)
```

# Creating a sf data frame from an aspatial data frame

The code chunk below converts `sgpools` data frame into a simple feature data frame by using `st_as_sf()` of sf packages.

```{r}
sgpools_sf <- st_as_sf(sgpools,
                       coords = c("XCOORD", "YCOORD"),
                       crs= 3414)
```

::: {.callout-note}
Things to note from the arguments above:

- The `coords` argument requires you to provide the column name of the x-coordinates first then followed by the column name of the y-coordinates.
- The `crs` argument requires you to provide the coordinates system in EPSG format. EPSG: 3414 is Singapore SVY21 Projected Coordinate System. You can search for other country's EPSG code by referring to [epsg.io](https://epsg.io).
:::

You can display the basic information of the newly created `sgpools_sf` by using the code chunk below.

```{r}
list(sgpools_sf)
```

The output shows that `sgpools_sf` is in point feature class. Its EPSG ID is 3414. The bbox provides information on the extent of the geospatial data.

# Drawing Proportional Symbol Map

To create an interactive proportional symbol map in R, the view mode of tmap will be used.

The code chunk below will turn on the interactive mode of tmap.

```{r}
tmap_mode("view")
```

## It all started with an interactive point symbol map

```{r}
tm_shape(sgpools_sf) +
  tm_bubbles(fill = "red",
           size = 1,
           col = "black",
           lwd = 1)
```

## Making it proportional

To draw a proportional symbol map, we need to assign a numerical variable to the size visual attribute. The code chunk below shows that the variable `Gp1Gp2 Winnings` is assigned to the size visual attribute.

```{r}
tm_shape(sgpools_sf) +
  tm_bubbles(fill = "red",
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1)
```

## Adding colour by outlet type

The proportional symbol map can be further improved by using the colour visual attribute. In the code chunk below, the `OUTLET TYPE` variable is used as the colour attribute variable.

```{r}
tm_shape(sgpools_sf) +
  tm_bubbles(fill = "OUTLET TYPE",
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1)
```

## Faceted Proportional Symbol Maps

An impressive and little-known feature of tmap's view mode is that it also works with faceted plots. The `sync` argument in `tm_facets()` can be used in this case to produce multiple maps with synchronised zoom and pan settings.

```{r}
tm_shape(sgpools_sf) +
  tm_bubbles(fill = "OUTLET TYPE",
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1) +
  tm_facets(by= "OUTLET TYPE",
            nrow = 1,
            sync = TRUE)
```

Before you end the session, it is wiser to switch tmap's Viewer back to plot mode by using the code chunk below.

```{r}
tmap_mode("plot")
```

# Reference
