---
title: "Hands-on Exercise 5 (Part 3)"
author: "Bui Tien Thanh (tt.bui.2025@mitb.smu.edu.sg)"
date: "2026-05-13"
date-modified: "last-modified"
description: "Heatmap for Visualising and Analysing Multivariate Data"
format: html
execute:
    echo: true
    eval: true
    warning: false
    freeze: true
---

# Heatmap for Visualising and Analysing Multivariate Data

## Overview

Heatmaps visualise data through variations in colouring. When applied to a tabular format, heatmaps are useful for cross-examining multivariate data, through placing variables in the columns and observation (or records) in rows and colouring the cells within the table. Heatmaps are good for showing variance across multiple variables, revealing any patterns, displaying whether any variables are similar to each other, and for detecting if any correlations exist in-between them.

In this hands-on exercise, you will gain hands-on experience on using R to plot static and interactive heatmap for visualising and analysing multivariate data.

## Installing and Launching R Packages

```{r}
pacman::p_load(seriation, dendextend, heatmaply, tidyverse)
```

## Data Preparation

### The data

In this hands-on exercise, the data of World Happiness 2018 report will be used.

### Importing Data

```{r}
wh <- read_csv("../data/WHData-2018.csv")
```

### Preparing the Data

```{r}
row.names(wh) <- wh$Country
```

### Transforming the data frame into a matrix

The data was loaded into a data frame, but it has to be a data matrix to make your heatmap.

```{r}
wh1 <- dplyr::select(wh, c(3, 7:12))
wh_matrix <- data.matrix(wh)
```

Notice that `wh_matrix` is in R matrix format.

## Static Heatmap

### heatmap() of R Stats

```{r}
wh_heatmap <- heatmap(wh_matrix,
                      Rowv=NA, Colv=NA)
```

::: callout-note
By default, `heatmap()` plots a cluster heatmap. The arguments `Rowv=NA` and `Colv=NA` are used to switch off the option of plotting the row and column dendrograms. To plot a cluster heatmap, we just have to use the default as shown in the code chunk below.
:::

```{r}
wh_heatmap <- heatmap(wh_matrix)
```

::: callout-note
The order of both rows and columns is different compared to the native `wh_matrix`. This is because heatmap does a reordering using clusterisation: it calculates the distance between each pair of rows and columns and tries to order them by similarity. Moreover, the corresponding dendrogram are provided beside the heatmap.

Here, red cells denote small values. This heatmap is not really informative because the Happiness Score variable has relatively higher values, making the other variables with small values all look the same. Thus, we need to normalize this matrix using the `scale` argument.
:::

The code chunk below normalises the matrix column-wise.

```{r}
wh_heatmap <- heatmap(wh_matrix,
                      scale="column",
                      cexRow = 0.6,
                      cexCol = 0.8,
                      margins = c(10, 4))
```

Notice that the values are scaled now. Also note that `margins` argument is used to ensure that the entire x-axis labels are displayed completely and, `cexRow` and `cexCol` arguments are used to define the font size used for y-axis and x-axis labels respectively.

## Creating Interactive Heatmap

heatmaply is an R package for building interactive cluster heatmap that can be shared online as a stand-alone HTML file. It is designed and maintained by Tal Galili.

In this section, you will gain hands-on experience on using heatmaply to design an interactive cluster heatmap. We will still use the `wh_matrix` as the input data.

### Working with heatmaply

```{r}
heatmaply(mtcars)
```

The code chunk below shows the basic syntax needed to create an interactive heatmap by using heatmaply package.

```{r}
heatmaply(wh_matrix[, -c(1, 2, 4, 5)])
```

::: callout-note
- Different from `heatmap()`, for `heatmaply()` the default horizontal dendrogram is placed on the left hand side of the heatmap.
- The text label of each row, on the other hand, is placed on the right hand side of the heat map.
- When the x-axis marker labels are too long, they will be rotated by 135 degree from the north.
:::

### Data transformation

When analysing multivariate data set, it is very common that the variables in the data sets includes values that reflect different types of measurement. In order to ensure that all the variables have comparable values, data transformation are commonly used before clustering.

Three main data transformation methods are supported by `heatmaply()`, namely: scale, normalise and percentilse.

#### Scaling method

When all variables are came from or assumed to come from some normal distribution, then scaling (i.e.: subtract the mean and divide by the standard deviation) would bring them all close to the standard normal distribution. In such a case, each value would reflect the distance from the mean in units of standard deviation. The `scale` argument in `heatmaply()` supports column and row scaling.

```{r}
heatmaply(wh_matrix[, -c(1, 2, 4, 5)],
          scale = "column")
```

#### Normalising method

When variables in the data comes from possibly different (and non-normal) distributions, the normalize function can be used to bring data to the 0 to 1 scale by subtracting the minimum and dividing by the maximum of all observations. This preserves the shape of each variable's distribution while making them easily comparable on the same "scale". Different from Scaling, the normalise method is performed on the input data set i.e. `wh_matrix` as shown in the code chunk below.

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]))
```

#### Percentising method

This is similar to ranking the variables, but instead of keeping the rank values, divide them by the maximal rank. This is done by using the ecdf of the variables on their own values, bringing each value to its empirical percentile. The benefit of the percentize function is that each value has a relatively clear interpretation, it is the percent of observations that got that value or below it. Similar to Normalize method, the Percentize method is also performed on the input data set i.e. `wh_matrix` as shown in the code chunk below.

```{r}
heatmaply(percentize(wh_matrix[, -c(1, 2, 4, 5)]))
```

### Clustering algorithm

heatmaply supports a variety of hierarchical clustering algorithm. The main arguments provided are:

- `distfun`: function used to compute the distance (dissimilarity) between both rows and columns. Defaults to `dist`. The options `"pearson"`, `"spearman"` and `"kendall"` can be used to use correlation-based clustering, which uses `as.dist(1 - cor(t(x)))` as the distance metric.
- `hclustfun`: function used to compute the hierarchical clustering when `Rowv` or `Colv` are not dendrograms. Defaults to `hclust`.
- `dist_method`: default is NULL, which results in `"euclidean"` to be used. It can accept alternative character strings indicating the method to be passed to `distfun`. By default `distfun` is `"dist"` hence this can be one of `"euclidean"`, `"maximum"`, `"manhattan"`, `"canberra"`, `"binary"` or `"minkowski"`.
- `hclust_method`: default is NULL, which results in `"complete"` method to be used. It can accept alternative character strings indicating the method to be passed to `hclustfun`. By default `hclustfun` is `hclust` hence this can be one of `"ward.D"`, `"ward.D2"`, `"single"`, `"complete"`, `"average"` (= UPGMA), `"mcquitty"` (= WPGMA), `"median"` (= WPGMC) or `"centroid"` (= UPGMC).

In general, a clustering model can be calibrated either manually or statistically.

### Manual approach

In the code chunk below, the heatmap is plotted by using hierarchical clustering algorithm with "Euclidean distance" and "ward.D" method.

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          dist_method = "euclidean",
          hclust_method = "ward.D")
```

### Statistical approach

In order to determine the best clustering method and number of cluster the `dend_expend()` and `find_k()` functions of dendextend package will be used.

First, the `dend_expend()` will be used to determine the recommended clustering method to be used.

```{r}
wh_d <- dist(normalize(wh_matrix[, -c(1, 2, 4, 5)]), method = "euclidean")
dend_expend(wh_d)[[3]]
```

The output table shows that "average" method should be used because it gave the high optimum value.

Next, `find_k()` is used to determine the optimal number of cluster.

```{r}
wh_clust <- hclust(wh_d, method = "average")
num_k <- find_k(wh_clust)
plot(num_k)
```

Figure above shows that k=3 would be good.

With reference to the statistical analysis results, we can prepare the code chunk as shown below.

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          dist_method = "euclidean",
          hclust_method = "average",
          k_row = 3)
```

### Seriation

One of the problems with hierarchical clustering is that it doesn't actually place the rows in a definite order, it merely constrains the space of possible orderings. heatmaply uses the seriation package to find an optimal ordering of rows and columns. Optimal means to optimize the Hamiltonian path length that is restricted by the dendrogram structure. This, in other words, means to rotate the branches so that the sum of distances between each adjacent leaf (label) will be minimized.

Here we meet our first seriation algorithm: Optimal Leaf Ordering (OLO). This algorithm starts with the output of an agglomerative clustering algorithm and produces a unique ordering, one that flips the various branches of the dendrogram around so as to minimize the sum of dissimilarities between adjacent leaves.

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          seriate = "OLO")
```

The default options is `"OLO"` (Optimal leaf ordering) which optimizes the above criterion (in O(n^4)). Another option is `"GW"` (Gruvaeus and Wainer) which aims for the same goal but uses a potentially faster heuristic.

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          seriate = "GW")
```

The option `"mean"` gives the output we would get by default from heatmap functions in other packages such as `gplots::heatmap.2`.

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          seriate = "mean")
```

The option `"none"` gives us the dendrograms without any rotation that is based on the data matrix.

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          seriate = "none")
```

### Working with colour palettes

The default colour palette used by heatmaply is viridis. heatmaply users, however, can use other colour palettes in order to improve the aestheticness and visual friendliness of the heatmap.

In the code chunk below, the Blues colour palette of rColorBrewer is used.

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          seriate = "none",
          colors = Blues)
```

### The finishing touch

Beside providing a wide collection of arguments for meeting the statistical analysis needs, heatmaply also provides many plotting features to ensure cartographic quality heatmap can be produced.

In the code chunk below the following arguments are used:

- `k_row` is used to produce 5 groups.
- `margins` is used to change the top margin to 60 and row margin to 200.
- `fontsize_row` and `fontsize_col` are used to change the font size for row and column labels to 4.
- `main` is used to write the main title of the plot.
- `xlab` and `ylab` are used to write the x-axis and y-axis labels respectively.

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          Colv=NA,
          seriate = "none",
          colors = Blues,
          k_row = 5,
          margins = c(NA,200,60,NA),
          fontsize_row = 4,
          fontsize_col = 5,
          main="World Happiness Score and Variables by Country, 2018 \nData Transformation using Normalise Method",
          xlab = "World Happiness Indicators",
          ylab = "World Countries"
          )
```
