---
title: "Hands-on Exercise 5 (Part 2)"
author: "Bui Tien Thanh (tt.bui.2025@mitb.smu.edu.sg)"
date: "2026-05-13"
date-modified: "last-modified"
description: "Creating Ternary Plot with R"
format: html
execute:
    echo: true
    eval: true
    warning: false
    freeze: true
---

# Creating Ternary Plot with R

## Overview

Ternary plots are a way of displaying the distribution and variability of three-part compositional data. (For example, the proportion of aged, economy active and young population or sand, silt, and clay in soil.) It’s display is a triangle with sides scaled from 0 to 1. Each side represents one of the three components. A point is plotted so that a line drawn perpendicular from the point to each leg of the triangle intersect at the component values of the point.

In this hands-on, you will learn how to build ternary plot programmatically using R for visualising and analysing population structure of Singapore.

The hands-on exercise consists of four steps:

1. Install and launch tidyverse and ggtern packages.
2. Derive three new measures using `mutate()` function of dplyr package.
3. Build a static ternary plot using `ggtern()` function of ggtern package.
4. Build an interactive ternary plot using `plot_ly()` function of Plotly R package.

## Installing and Launching R Packages

```{r}
pacman::p_load(plotly, ggtern, tidyverse)
```

## Data Preparation

### The data

For the purpose of this hands-on exercise, the Singapore Residents by Planning Area/Subzone, Age Group, Sex and Type of Dwelling, June 2000-2018 data will be used.

### Importing Data

```{r}
pop_data <- read_csv("../data/respopagsex2000to2018_tidy.csv")
```

### Preparing the Data

```{r}
agpop_mutated <- pop_data %>%
  mutate(`Year` = as.character(Year))%>%
  spread(AG, Population) %>%
  mutate(YOUNG = rowSums(.[4:8]))%>%
  mutate(ACTIVE = rowSums(.[9:16]))  %>%
  mutate(OLD = rowSums(.[17:21])) %>%
  mutate(TOTAL = rowSums(.[22:24])) %>%
  filter(Year == 2018)%>%
  filter(TOTAL > 0)
```

## Plotting Ternary Diagram with R

### Plotting a static ternary diagram

Use `ggtern()` function of ggtern package to create a simple ternary plot.

```{r}
ggtern(data=agpop_mutated,aes(x=YOUNG,y=ACTIVE, z=OLD)) +
  geom_point()
```

```{r}
ggtern(data=agpop_mutated, aes(x=YOUNG,y=ACTIVE, z=OLD)) +
  geom_point() +
  labs(title="Population structure, 2015") +
  theme_rgbw()
```

### Plotting an interactive ternary diagram

The code below create an interactive ternary plot using plot_ly() function of Plotly R.

```{r}
# reusable function for creating annotation object
label <- function(txt) {
  list(
    text = txt, 
    x = 0.1, y = 1,
    ax = 0, ay = 0,
    xref = "paper", yref = "paper", 
    align = "center",
    font = list(family = "serif", size = 15, color = "white"),
    bgcolor = "#b3b3b3", bordercolor = "black", borderwidth = 2
  )
}

# reusable function for axis formatting
axis <- function(txt) {
  list(
    title = txt, tickformat = ".0%", tickfont = list(size = 10)
  )
}

ternaryAxes <- list(
  aaxis = axis("Young"), 
  baxis = axis("Active"), 
  caxis = axis("Old")
)

# Initiating a plotly visualization 
plot_ly(
  agpop_mutated, 
  a = ~YOUNG, 
  b = ~ACTIVE, 
  c = ~OLD, 
  color = I("black"), 
  type = "scatterternary"
) %>%
  layout(
    annotations = label("Ternary Markers"), 
    ternary = ternaryAxes
  )
```

