---
title: "Hands-on Exercise 4 (Part 4)"
author: "Bui Tien Thanh (tt.bui.2025@mitb.smu.edu.sg)"
date: "2026-05-07"
date-modified: "last-modified"
description: "Visualising Uncertainty with R"
format: html
execute:
    echo: true
    eval: true
    warning: false
    freeze: true
---

# Visualising Uncertainty

## Learning Outcome

In this hands-on exercise, you will learn how to visualize uncertainty using various R packages and techniques.

## Getting Started

### Installing and Loading the Packages

```{r}
pacman::p_load(plotly, crosstalk, DT,
               ggdist, ggridges, colorspace,
               gganimate, tidyverse)
```

### Data Import

```{r}
exam <- read_csv("../data/Exam_data.csv")
```

## Visualizing the Uncertainty of Point Estimates: ggplot2 Methods

A point estimate is a single number, such as a mean. Uncertainty, on the other hand, is expressed as standard error, confidence interval, or credible interval.

Firstly, code chunk below will be used to derive the necessary summary statistics.

```{r}
my_sum <- exam %>%
  group_by(RACE) %>%
  summarise(
    n=n(),
    mean=mean(MATHS),
    sd=sd(MATHS)
  ) %>%
  mutate(se=sd/sqrt(n-1))
```

Next, the code chunk below will be used to display my_sum tibble data frame in an html table format.

```{r}
knitr::kable(head(my_sum), format = 'html')
```

### Plotting Standard Error Bars of Point Estimates

Now we are ready to plot the standard error bars of mean maths score by race as shown below.

```{r}
ggplot(my_sum) +
  geom_errorbar(
    aes(x=RACE,
        ymin=mean-se,
        ymax=mean+se),
    width=0.2,
    colour="black",
    alpha=0.9,
    linewidth=0.5) +
  geom_point(aes(x=RACE,
            y=mean),
           stat="identity",
           color="red",
           size = 1.5,
           alpha=1) +
  ggtitle("Standard error of mean maths score by race")
```

### Plotting Confidence Interval of Point Estimates

Instead of plotting the standard error bar of point estimates, we can also plot the confidence intervals of mean maths score by race.

```{r}
ggplot(my_sum) +
  geom_errorbar(
    aes(x=reorder(RACE, -mean),
        ymin=mean-1.96*se,
        ymax=mean+1.96*se),
    width=0.2,
    colour="black",
    alpha=0.9,
    linewidth=0.5) +
  geom_point(aes(x=RACE,
            y=mean),
           stat="identity",
           color="red",
           size = 1.5,
           alpha=1) +
  labs(x = "Maths score",
       title = "95% confidence interval of mean maths score by race")
```

### Visualizing the Uncertainty of Point Estimates with Interactive Error Bars

In this section, you will learn how to plot interactive error bars for the 99% confidence interval of mean maths score by race as shown in the figure below.

```{r}
shared_df = SharedData$new(my_sum)

bscols(widths = c(4,8),
       ggplotly((ggplot(shared_df) +
                  geom_errorbar(aes(
                    x=reorder(RACE, -mean),
                    ymin=mean-2.58*se,
                    ymax=mean+2.58*se),
                    width=0.2,
                    colour="black",
                    alpha=0.9,
                    size=0.5) +
                  geom_point(aes(
                    x=RACE,
                    y=mean,
                    text = paste("Race:", `RACE`,
                                 "<br>N:", `n`,
                                 "<br>Avg. Scores:", round(mean, digits = 2),
                                 "<br>95% CI:[",
                                 round((mean-2.58*se), digits = 2), ",",
                                 round((mean+2.58*se), digits = 2),"]")),
                    stat="identity",
                    color="red",
                    size = 1.5,
                    alpha=1) +
                  xlab("Race") +
                  ylab("Average Scores") +
                  theme_minimal() +
                  theme(axis.text.x = element_text(
                    angle = 45, vjust = 0.5, hjust=1)) +
                  ggtitle("99% Confidence interval of average /<br>maths scores by race")),
                tooltip = "text"),
       DT::datatable(shared_df,
                     rownames = FALSE,
                     class="compact",
                     width="100%",
                     options = list(pageLength = 10,
                                    scrollX=T),
                     colnames = c("No. of pupils",
                                  "Avg Scores",
                                  "Std Dev",
                                  "Std Error")) %>%
         formatRound(columns=c('mean', 'sd', 'se'),
                     digits=2))
```

## Visualising Uncertainty: ggdist Package

ggdist is an R package that provides a flexible set of ggplot2 geoms and stats designed especially for visualising distributions and uncertainty.

It is designed for both frequentist and Bayesian uncertainty visualization, taking the view that uncertainty visualization can be unified through the perspective of distribution visualization:

- For frequentist models, one visualises confidence distributions or bootstrap distributions (see vignette(“freq-uncertainty-vis”))
- For Bayesian models, one visualises probability distributions (see the tidybayes package, which builds on top of ggdist)

### Visualizing the Uncertainty of Point Estimates: ggdist Methods

In the code chunk below, stat_pointinterval() of ggdist is used to build a visual for displaying distribution of maths scores by race.

```{r}
exam %>%
  ggplot(aes(x = RACE,
             y = MATHS)) +
  stat_pointinterval() +
  labs(
    title = "Visualising confidence intervals of mean math score",
    subtitle = "Mean Point + Multiple-interval plot")
```

For example, in the code chunk below the following arguments are used:

- `.width = 0.95`
- `.point = median`
- `.interval = qi`

```{r}
exam %>%
  ggplot(aes(x = RACE, y = MATHS)) +
  stat_pointinterval(.width = 0.95,
  .point = median,
  .interval = qi) +
  labs(
    title = "Visualising confidence intervals of median math score",
    subtitle = "Median Point + Multiple-interval plot")
```

### Visualizing Distribution with stat_pointinterval()

```{r}
exam %>%
  ggplot(aes(x = RACE,
             y = MATHS)) +
  stat_pointinterval(
    show.legend = FALSE) +
  labs(
    title = "Visualising confidence intervals of mean math score",
    subtitle = "Mean Point + Multiple-interval plot")
```

### Visualizing Distribution with stat_gradientinterval()

In the code chunk below, stat_gradientinterval() of ggdist is used to build a visual for displaying distribution of maths scores by race.

```{r}
exam %>%
  ggplot(aes(x = RACE,
             y = MATHS)) +
  stat_gradientinterval(
    fill = "skyblue",
    show.legend = TRUE
  ) +
  labs(
    title = "Visualising confidence intervals of mean math score",
    subtitle = "Gradient + interval plot")
```

## Visualising Uncertainty with Hypothetical Outcome Plots (HOPs)

### Installing ungeviz Package

```{r}
devtools::install_github("wilkelab/ungeviz")
```

### Launch the Application in R

```{r}
library(ungeviz)
```

### Visualising Uncertainty with Hypothetical Outcome Plots (HOPs)

Next, the code chunk below will be used to build the HOPs.

```{r}
ggplot(data = exam,
       (aes(x = factor(RACE),
            y = MATHS))) +
  geom_point(position = position_jitter(
    height = 0.3,
    width = 0.05),
    size = 0.4,
    color = "#0072B2",
    alpha = 1/2) +
  geom_hpline(data = sampler(25,
                             group = RACE),
              height = 0.6,
              color = "#D55E00") +
  theme_bw() +
  transition_states(.draw, 1, 3)
```

