---
title: "Take-home Exercise 2: VAST Challenge 2026 MC1"
author: "Bui Tien Thanh"
date: "May 9, 2026"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
---

## Overview

### Background

This analysis investigates the **TenantThread embargo breach incident** that occurred on June 5th, 2046. TenantThread, a property tech company, had signed a merger agreement with CivicLoom Realty Partners (codenamed "Project HarborCrest"). The merger announcement was under strict embargo until 6:00 PM, but embargoed information leaked on social media around 5:00 PM.

### Main Objective

**Central Question:** Did TenantThread's team deliberately leak the merger or did the system simply break down under pressure?

This analysis examines internal communications from 7 AI agents over two weeks leading up to the breach to answer this question.

### Analytical Approach

- **Task 1 (Essential):** Event Sequence Visualization - Reconstruct the chain of events leading to the breach
- **Task 2 (Essential):** Behavioral Analysis - Compare breach behavior to baseline patterns to determine intent vs. accident

## Loading Packages

```{r}
pacman::p_load(
  tidyverse,      # Data manipulation and visualization
  jsonlite,       # JSON data handling
  lubridate,      # Date/time manipulation
  igraph,         # Network analysis
  visNetwork,     # Interactive network visualization
  plotly,         # Interactive plots
  DT,             # Interactive tables
  tidytext,       # Text analysis
  scales,         # Scale functions for visualization
  ggtext          # Enhanced text rendering
)
```

## Style Configuration

### Define Styles

```{r}
STYLES <- list(
  # Font
  font_family = "sans",

  # Agent colors (colorblind-friendly palette)
  agent_colors = c(
    "judge_eval_agent" = "#D81B60",      # Pink-red (Judge/Compliance)
    "legal_agent" = "#1E88E5",           # Blue (Legal)
    "pr_agent" = "#FFC107",              # Amber (PR)
    "pr_intern_agent" = "#FFE082",       # Light amber (PR Intern)
    "quality_agent" = "#004D40",         # Teal (Quality)
    "social_manager_agent" = "#7B1FA2",  # Purple (Social)
    "intern_agent" = "#9E9E9E"           # Grey (Intern)
  ),

  # Period colors
  period_colors = c(
    "Baseline (Week 1)" = "#66BB6A",     # Green
    "Final Week" = "#FFA726",            # Orange
    "Breach Day" = "#EF5350"             # Red
  ),

  # Action colors
  action_colors = c(
    "MONITORING" = "#90CAF9",
    "ESCALATING" = "#FFAB91",
    "RESPONDING" = "#CE93D8",
    "PUBLISHING" = "#80CBC4"
  ),

  # Plot elements
  title_color = "grey25",
  text_color = "grey50",
  panel_border_color = "grey50",

  # Interactive
  tooltip_css = "background-color:black;color:white;font-size:10pt;padding:4px;"
)
```

### Define Common Theme

```{r}
COMMON_THEME <- theme_minimal() +
  theme(
    text = element_text(family = STYLES$font_family, color = STYLES$text_color),
    plot.title = element_text(
      color = STYLES$title_color,
      face = "bold",
      size = 12
    ),
    plot.subtitle = element_text(size = 10),
    plot.caption = element_text(size = 8, hjust = 0),

    legend.position = "right",
    legend.title = element_text(face = "bold", size = 10),
    legend.text = element_text(size = 9),

    panel.border = element_rect(
      color = STYLES$panel_border_color,
      fill = NA,
      linewidth = 0.5
    ),

    axis.text = element_text(size = 9),
    axis.title = element_text(size = 10)
  )
```

## Data Preparation

### Loading the Data

```{r}
# Load MC1 data
mc1_data <- fromJSON("../../data/MC1_final.json")

# Extract rounds
rounds <- mc1_data$rounds
```

### Data Structure Exploration

```{r}
# Check number of rounds (hours)
cat("Total rounds:", length(rounds), "\n")

# Examine first round structure
cat("\nFirst round timestamp:", rounds[[1]]$hour, "\n")
cat("Number of agents in first round:", length(rounds[[1]]$agent_outputs), "\n")
```

### Extract and Transform Data

```{r}
# Create a dataframe of all agent communications
communications_df <- map_dfr(1:length(rounds), function(i) {
  round <- rounds[[i]]
  hour <- round$hour

  map_dfr(round$agent_outputs, function(agent) {
    if (length(agent$communications) > 0) {
      map_dfr(agent$communications, function(comm) {
        tibble(
          hour = hour,
          agent_id = agent$agent_id,
          message_id = comm$message_id,
          channel = comm$channel,
          message_text = comm$message_text,
          declared_action = agent$declared_action
        )
      })
    }
  })
})

# Parse datetime
communications_df <- communications_df %>%
  mutate(datetime = ymd_h(hour))

# Display summary
cat("Total communications:", nrow(communications_df), "\n")
```

### Extract Environment Context

```{r}
# Extract environment context to understand external pressures
environment_df <- map_dfr(1:length(rounds), function(i) {
  round <- rounds[[i]]
  tibble(
    hour = round$hour,
    event_narrative = round$environment_context$event_narrative %||% NA_character_,
    agents_unavailable = paste(unlist(round$environment_context$agents_unavailable), collapse = ", ")
  )
}) %>%
  mutate(datetime = ymd_h(hour))

cat("Environment events tracked:", nrow(environment_df), "\n")
```

### Extract Complete Internal States

```{r}
# Extract internal states for all agents (already done above but moved here for clarity)
internal_states_df <- map_dfr(1:length(rounds), function(i) {
  round <- rounds[[i]]
  hour <- round$hour

  map_dfr(round$agent_outputs, function(agent) {
    tibble(
      hour = hour,
      agent_id = agent$agent_id,
      reacting = agent$internal_state$reacting %||% NA_character_,
      rationalizing = agent$internal_state$rationalizing %||% NA_character_,
      deliberating = agent$internal_state$deliberating %||% NA_character_,
      declared_action = agent$declared_action
    )
  })
}) %>%
  mutate(datetime = ymd_h(hour))

cat("Internal state records:", nrow(internal_states_df), "\n")
```

### Build Agent Interaction Network

```{r}
# Create agent interaction data (who responds to whom)
agent_interactions <- communications_df %>%
  arrange(datetime, agent_id) %>%
  group_by(channel) %>%
  mutate(
    previous_agent = lag(agent_id),
    previous_message = lag(message_text),
    time_since_last = as.numeric(difftime(datetime, lag(datetime), units = "mins"))
  ) %>%
  filter(!is.na(previous_agent), previous_agent != agent_id, time_since_last <= 60) %>%
  ungroup()

cat("Agent interactions identified:", nrow(agent_interactions), "\n")
```

## Analysis

### Exploratory Data Analysis

#### Agent Activity Overview

```{r}
# Count communications by agent
agent_activity <- communications_df %>%
  count(agent_id, name = "total_messages") %>%
  arrange(desc(total_messages))

# Visualize
ggplot(agent_activity, aes(x = reorder(agent_id, total_messages), y = total_messages, fill = agent_id)) +
  geom_col() +
  scale_fill_manual(values = STYLES$agent_colors) +
  coord_flip() +
  labs(title = "Total Communications by Agent",
       x = "Agent", y = "Number of Messages") +
  COMMON_THEME +
  theme(legend.position = "none")
```

#### Communication Channels

```{r}
# Channel usage
channel_usage <- communications_df %>%
  count(channel, name = "messages") %>%
  arrange(desc(messages))

datatable(channel_usage, options = list(pageLength = 10))
```

#### Temporal Pattern

```{r}
# Messages over time
temporal_pattern <- communications_df %>%
  count(datetime, name = "messages")

ggplot(temporal_pattern, aes(x = datetime, y = messages)) +
  geom_line(color = STYLES$agent_colors["legal_agent"], linewidth = 1) +
  geom_point(color = STYLES$agent_colors["legal_agent"], size = 2) +
  labs(title = "Communication Volume Over Time",
       x = "Date", y = "Number of Messages") +
  COMMON_THEME +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Task 1: Event Sequence Visualization ⭐

#### Identifying the Breach Event

```{r}
# Find communications on breach day (June 5, 2046)
breach_day <- communications_df %>%
  filter(date(datetime) == ymd("2046-06-05")) %>%
  arrange(datetime)

# Focus on critical hours (4 PM - 6 PM, embargo at 6 PM, breach at 5 PM)
breach_window <- breach_day %>%
  filter(hour(datetime) >= 16, hour(datetime) <= 18)

cat("Messages during breach window (4-6 PM):", nrow(breach_window), "\n")
cat("Agents active during breach:", n_distinct(breach_window$agent_id), "\n")

# Identify the actual breach message(s)
breach_messages <- breach_window %>%
  filter(contains_embargo_terms, hour(datetime) == 17) %>%
  arrange(datetime)

cat("Potential breach messages at 5 PM:", nrow(breach_messages), "\n")
```

#### Breach Event Details

```{r}
# Show breach messages
if(nrow(breach_messages) > 0) {
  breach_details <- breach_messages %>%
    select(datetime, agent_id, channel, message_text) %>%
    mutate(message_text = str_trunc(message_text, 100))

  datatable(breach_details,
            caption = "Messages with Embargo Terms at 5 PM (Breach Hour)",
            options = list(pageLength = 5))
}
```

#### Causal Chain Analysis

```{r}
# Analyze agent interactions during breach window
breach_interactions <- agent_interactions %>%
  filter(date(datetime) == ymd("2046-06-05"), hour(datetime) >= 16) %>%
  select(datetime, previous_agent, agent_id, channel, time_since_last, message_text) %>%
  arrange(datetime)

# Create interaction flow
interaction_flow <- breach_interactions %>%
  count(previous_agent, agent_id, channel) %>%
  arrange(desc(n))

datatable(interaction_flow,
          caption = "Agent Interaction Flow During Breach Window",
          options = list(pageLength = 10))
```

#### Event Sequence Timeline

```{r}
# Build timeline of key events on breach day
breach_timeline <- internal_states_df %>%
  filter(date(datetime) == ymd("2046-06-05")) %>%
  left_join(
    communications_df %>%
      group_by(datetime, agent_id) %>%
      summarise(num_messages = n(), .groups = "drop"),
    by = c("datetime", "agent_id")
  ) %>%
  replace_na(list(num_messages = 0)) %>%
  arrange(datetime, agent_id)

# Visualize agent activity timeline
ggplot(breach_timeline, aes(x = datetime, y = agent_id)) +
  geom_point(aes(size = num_messages, color = declared_action), alpha = 0.7) +
  scale_color_manual(values = STYLES$action_colors, na.value = "grey70") +
  geom_vline(xintercept = ymd_h("2046-06-05 17"), linetype = "dashed",
             color = STYLES$period_colors["Breach Day"], linewidth = 1) +
  annotate("text", x = ymd_h("2046-06-05 17"), y = 7, label = "Breach (5 PM)",
           hjust = -0.1, color = STYLES$period_colors["Breach Day"], size = 3) +
  geom_vline(xintercept = ymd_h("2046-06-05 18"), linetype = "dashed",
             color = STYLES$agent_colors["legal_agent"], linewidth = 1) +
  annotate("text", x = ymd_h("2046-06-05 18"), y = 7, label = "Embargo End (6 PM)",
           hjust = -0.1, color = STYLES$agent_colors["legal_agent"], size = 3) +
  labs(title = "Agent Activity Timeline on Breach Day",
       x = "Time", y = "Agent", size = "Messages", color = "Action") +
  COMMON_THEME +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


#### Interactive Communication Network (Breach Window)

```{r}
# Build network for breach window only
breach_network_edges <- agent_interactions %>%
  filter(date(datetime) == ymd("2046-06-05"), hour(datetime) >= 16) %>%
  count(previous_agent, agent_id, name = "weight")

# Create nodes
breach_network_nodes <- data.frame(
  id = unique(c(breach_network_edges$previous_agent, breach_network_edges$agent_id)),
  label = unique(c(breach_network_edges$previous_agent, breach_network_edges$agent_id)),
  group = c("judge_eval_agent", "legal_agent", "pr_agent", "pr_intern_agent",
            "quality_agent", "social_manager_agent", "intern_agent")
)

# Interactive network
visNetwork(breach_network_nodes,
           breach_network_edges %>% rename(from = previous_agent, to = agent_id),
           main = "Agent Interaction Network During Breach Window (4-6 PM)") %>%
  visEdges(arrows = "to") %>%
  visOptions(highlightNearest = TRUE, nodesIdSelection = TRUE) %>%
  visLayout(randomSeed = 123)
```

#### Judge Agent's Role During Breach

```{r}
# Extract Judge agent activity during breach day
judge_breach_day <- internal_states_df %>%
  filter(agent_id == "judge_eval_agent", date(datetime) == ymd("2046-06-05")) %>%
  arrange(datetime)

# Show Judge's internal states
judge_states_table <- judge_breach_day %>%
  select(datetime, declared_action, reacting, rationalizing, deliberating) %>%
  mutate(across(c(reacting, rationalizing, deliberating), ~str_trunc(., 80)))

datatable(judge_states_table,
          caption = "Judge Agent Internal States on Breach Day",
          options = list(pageLength = 5, scrollX = TRUE))
```


### Task 2: Behavioral Analysis ⭐

#### Define Analysis Periods

```{r}
# Define three periods for comparison
internal_states_df <- internal_states_df %>%
  mutate(
    period = case_when(
      datetime < ymd("2046-05-29") ~ "Baseline (Week 1)",
      datetime < ymd("2046-06-05") ~ "Final Week",
      date(datetime) == ymd("2046-06-05") ~ "Breach Day",
      TRUE ~ "Other"
    )
  )

# Period summary
period_summary <- internal_states_df %>%
  filter(period != "Other") %>%
  count(period) %>%
  arrange(period)

datatable(period_summary, caption = "Analysis Periods", options = list(dom = 't'))
```

#### Baseline Behavior Profiles

```{r}
# Establish baseline behavior for each agent (Week 1)
baseline_behavior <- internal_states_df %>%
  filter(period == "Baseline (Week 1)") %>%
  count(agent_id, declared_action) %>%
  group_by(agent_id) %>%
  mutate(baseline_pct = n / sum(n) * 100) %>%
  ungroup()

# Visualize baseline behavior by agent
ggplot(baseline_behavior, aes(x = agent_id, y = baseline_pct, fill = declared_action)) +
  geom_col() +
  scale_fill_manual(values = STYLES$action_colors, na.value = "grey70") +
  coord_flip() +
  labs(title = "Baseline Behavior Profile (Week 1)",
       x = "Agent", y = "Percentage (%)", fill = "Action") +
  COMMON_THEME
```


#### Behavioral Comparison Across Periods

```{r}
# Compare behavior across all three periods
period_behavior <- internal_states_df %>%
  filter(period != "Other") %>%
  count(period, agent_id, declared_action) %>%
  group_by(period, agent_id) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup()

# Visualize by period
ggplot(period_behavior, aes(x = agent_id, y = pct, fill = declared_action)) +
  geom_col() +
  scale_fill_manual(values = STYLES$action_colors, na.value = "grey70") +
  coord_flip() +
  facet_wrap(~period, ncol = 3) +
  labs(title = "Agent Behavior Comparison Across Periods",
       x = "Agent", y = "Percentage (%)", fill = "Action") +
  COMMON_THEME +
  theme(axis.text.x = element_text(size = 7))
```

#### Anomaly Detection

```{r}
# Identify unusual actions (actions that occur < 5% in baseline)
rare_actions <- baseline_behavior %>%
  filter(baseline_pct < 5) %>%
  select(agent_id, declared_action, baseline_pct)

# Find when these rare actions occurred in final week and breach day
anomalies <- internal_states_df %>%
  filter(period %in% c("Final Week", "Breach Day")) %>%
  inner_join(rare_actions, by = c("agent_id", "declared_action")) %>%
  select(datetime, period, agent_id, declared_action, baseline_pct) %>%
  arrange(datetime)

if(nrow(anomalies) > 0) {
  datatable(anomalies,
            caption = "Anomalous Actions (Rare in Baseline, Occurred Later)",
            options = list(pageLength = 10))
} else {
  cat("No significant anomalies detected based on baseline behavior.\n")
}
```


#### Internal State Analysis: Stress Indicators

```{r}
# Analyze internal states for stress/urgency keywords
stress_keywords <- c("urgent", "crisis", "pressure", "risk", "concern", "worried", "critical", "emergency")

internal_state_analysis <- internal_states_df %>%
  filter(period != "Other") %>%
  pivot_longer(
    cols = c(reacting, rationalizing, deliberating),
    names_to = "state_type",
    values_to = "state_content"
  ) %>%
  filter(!is.na(state_content)) %>%
  mutate(
    shows_stress = str_detect(tolower(state_content), paste(stress_keywords, collapse = "|")),
    mentions_judge = str_detect(tolower(state_content), "judge"),
    mentions_embargo = str_detect(tolower(state_content), "embargo|harborcrest|civicloom")
  )

# Count stress indicators by period
stress_by_period <- internal_state_analysis %>%
  group_by(period, agent_id) %>%
  summarise(
    total_states = n(),
    stress_count = sum(shows_stress),
    stress_pct = (stress_count / total_states) * 100,
    .groups = "drop"
  )

# Visualize stress levels
ggplot(stress_by_period, aes(x = agent_id, y = stress_pct, fill = period)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = STYLES$period_colors) +
  coord_flip() +
  labs(title = "Stress Indicators in Internal States by Period",
       x = "Agent", y = "% of States Showing Stress", fill = "Period") +
  COMMON_THEME
```


#### Communication Pattern Analysis

```{r}
# Analyze embargo-related communications
embargo_keywords <- c("merger", "CivicLoom", "HarborCrest", "acquisition", "deal")

communications_df <- communications_df %>%
  mutate(
    contains_embargo_terms = str_detect(
      tolower(message_text),
      paste(embargo_keywords, collapse = "|")
    ),
    period = case_when(
      datetime < ymd("2046-05-29") ~ "Baseline (Week 1)",
      datetime < ymd("2046-06-05") ~ "Final Week",
      date(datetime) == ymd("2046-06-05") ~ "Breach Day",
      TRUE ~ "Other"
    )
  )

# Embargo mentions by period
embargo_by_period <- communications_df %>%
  filter(period != "Other") %>%
  group_by(period, agent_id) %>%
  summarise(
    total_messages = n(),
    embargo_messages = sum(contains_embargo_terms),
    embargo_pct = (embargo_messages / total_messages) * 100,
    .groups = "drop"
  )

# Visualize
ggplot(embargo_by_period, aes(x = agent_id, y = embargo_pct, fill = period)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = STYLES$period_colors) +
  coord_flip() +
  labs(title = "Embargo-Related Communications by Period",
       x = "Agent", y = "% of Messages Mentioning Embargo", fill = "Period") +
  COMMON_THEME
```

#### Statistical Comparison: Baseline vs Breach Day

```{r}
# Compare key metrics
comparison_metrics <- internal_states_df %>%
  filter(period %in% c("Baseline (Week 1)", "Breach Day")) %>%
  group_by(period, agent_id) %>%
  summarise(
    total_actions = n(),
    unique_actions = n_distinct(declared_action),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = period,
    values_from = c(total_actions, unique_actions),
    names_sep = "_"
  )

datatable(comparison_metrics,
          caption = "Behavioral Metrics: Baseline vs Breach Day",
          options = list(pageLength = 10, scrollX = TRUE))
```


#### Judge Agent Compliance Analysis

```{r}
# Analyze Judge agent behavior across periods
judge_analysis <- internal_states_df %>%
  filter(agent_id == "judge_eval_agent", period != "Other") %>%
  count(period, declared_action) %>%
  group_by(period) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup()

# Visualize Judge's action distribution
ggplot(judge_analysis, aes(x = period, y = pct, fill = declared_action)) +
  geom_col() +
  scale_fill_manual(values = STYLES$action_colors, na.value = "grey70") +
  labs(title = "Judge Agent Action Distribution by Period",
       x = "Period", y = "Percentage (%)", fill = "Action") +
  COMMON_THEME +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#### Communication Channel Analysis

```{r}
# Analyze channel usage patterns across periods
channel_analysis <- communications_df %>%
  filter(period != "Other") %>%
  count(period, channel) %>%
  group_by(period) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup()

# Visualize
ggplot(channel_analysis, aes(x = channel, y = pct, fill = period)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = STYLES$period_colors) +
  coord_flip() +
  labs(title = "Communication Channel Usage by Period",
       x = "Channel", y = "Percentage (%)", fill = "Period") +
  COMMON_THEME
```


## Key Findings

### Evidence Summary

```{r}
# Create comprehensive summary table
findings_summary <- tibble(
  Metric = c(
    "Total Communications (2 weeks)",
    "Breach Day Communications",
    "Breach Window Messages (4-6 PM)",
    "Agents Involved",
    "Total Embargo-related Messages",
    "Embargo Messages on Breach Day",
    "Stress Indicators (Breach Day)",
    "Anomalous Actions Detected"
  ),
  Value = as.character(c(
    nrow(communications_df),
    nrow(breach_day),
    nrow(breach_window),
    n_distinct(communications_df$agent_id),
    sum(communications_df$contains_embargo_terms, na.rm = TRUE),
    sum(breach_day$contains_embargo_terms, na.rm = TRUE),
    nrow(internal_state_analysis %>% filter(period == "Breach Day", shows_stress)),
    nrow(anomalies)
  ))
)

datatable(findings_summary, options = list(dom = 't'))
```

## Conclusion

### Answer to Main Question

**Did TenantThread's team deliberately leak the merger or did the system simply break down under pressure?**

Based on the comprehensive visual analytics conducted, the evidence suggests:

#### Evidence Analysis

**1. Event Sequence (Task 1 Findings):**

- The breach occurred at 5 PM, exactly one hour before the embargo lifted at 6 PM
- Agent interaction network shows communication patterns during the breach window
- The Judge agent (compliance officer) was active but the system still failed
- Multiple agents were involved in communications during the critical timeframe

**2. Behavioral Patterns (Task 2 Findings):**

- **Baseline Comparison**: Agent behavior on breach day showed deviations from Week 1 baseline patterns
- **Stress Indicators**: Internal states revealed increased stress markers in the final week and breach day
- **Anomaly Detection**: Identified actions that were rare in baseline but occurred during critical periods
- **Communication Patterns**: Embargo-related communications increased significantly approaching breach day

**3. Judge Agent Role:**

- The Judge agent's compliance actions show activity throughout the period
- However, the breach still occurred despite the Judge's presence
- This suggests either:
  - The Judge's rules were insufficient to catch the specific breach scenario
  - The system was overwhelmed by the volume/complexity of decisions
  - There was a gap in the compliance logic

#### Conclusion

The evidence points toward **SYSTEM BREAKDOWN UNDER PRESSURE** rather than deliberate leak:

**Supporting Evidence:**
- Increased stress indicators across multiple agents
- Rising communication volume and embargo mentions over time (not sudden)
- Judge agent was active but failed to prevent breach (system limitation, not bypass)
- Behavioral changes consistent with pressure response rather than coordinated action
- No clear evidence of deliberate coordination to evade compliance

**Key Insight:** The breach appears to be a failure of the automated compliance system ("The Judge") to handle the complexity and pressure of the situation, rather than a deliberate attempt by the team to leak information.

### Recommendations

**1. System Improvements:**
- Enhance the Judge agent's decision-making logic to handle edge cases
- Implement multi-layer approval for embargo-sensitive communications
- Add stress detection and automatic escalation protocols

**2. Monitoring Enhancements:**
- Real-time monitoring of embargo-related communication frequency
- Alert system for unusual behavioral patterns
- Dashboard for compliance officer oversight

**3. Policy Changes:**
- Mandatory human review for all embargo-related posts within 24 hours of deadline
- Stricter channel restrictions during embargo periods
- Regular stress testing of compliance systems under pressure scenarios

## References

- VAST Challenge 2026 Mini Challenge 1 Dataset
- MC1 Data Description Documentation

