---
title: "Take-home Exercise 1 (Presentation): Motor Insurance Portfolio Analysis"
subtitle: "Visual Analytics of Profitability and Risk"
author: "Bui Tien Thanh"
date: "May 9, 2026"
format:
  revealjs:
    theme: simple
    transition: slide
    slide-number: true
    chalkboard: true
---

## Contents {.smaller}

**Setup**

1.  Introduction & Objectives
2.  Data Overview
3.  Portfolio Health Snapshot

**Key Drivers of Profitability**

4.  Customer Tenure Drives Profit
5.  Comprehensive Coverage Outperforms
6.  Location Matters
7.  Older Vehicles Carry Higher Risk
8.  Pricing Model Needs Enhancement

**Additional Insights**

9.  Claims Volatility by Segment
10. Fuel Type Shows Minimal Impact
11. Bonus Score Patterns

**Conclusions**

12. Key Insights Summary
13. Strategic Recommendations

::: notes
This presentation analyzes a Spanish motor insurance portfolio to identify what drives profitability and where risks lie. We examine over 600,000 policies to answer: Which customer segments are most profitable? Where should we focus pricing improvements? What risks need attention?
:::

## Introduction & Objectives

**Portfolio:** Spanish motor insurance with 600,000+ policies

**Business Questions:**

- Which customer segments drive profitability?
- Where are our pricing gaps and opportunities?
- What risks require immediate attention?

**Goal:** Identify actionable strategies to improve portfolio performance

::: notes
This analysis examines our motor insurance portfolio to find what makes policies profitable and where we're leaving money on the table. We look at customer characteristics, vehicle types, geographic patterns, and pricing effectiveness. The goal is to provide clear recommendations for pricing adjustments, risk management, and customer retention strategies.
:::

## Data Overview

```{r}
#| echo: false
#| message: false
pacman::p_load(tidyverse, ggplot2, readxl, ggstatsplot, scales, ggridges, ggdist, patchwork)

insurance_data <- read.csv(
  "../../data/Dataset of motor insurance portfolio.csv",
  sep = ";",
  dec = ".")

insurance_clean <- insurance_data %>%
  mutate(
    years_with_licence = year - age_driving_licence,
    profitability = total_premium - total_incurred,
    profit_ratio = ifelse(total_premium > 0, profitability / total_premium, NA),
    has_claim = ifelse(total_incurred > 0, "Yes", "No"),
    vehicle_age_group = case_when(
      vehicle_age < 0 ~ "Invalid",
      vehicle_age <= 5 ~ "0-5 years",
      vehicle_age <= 10 ~ "6-10 years",
      vehicle_age <= 20 ~ "11-20 years",
      vehicle_age <= 30 ~ "21-30 years",
      vehicle_age <= 40 ~ "31-40 years",
      vehicle_age <= 50 ~ "41-50 years",
      TRUE ~ "50+ years"
    ),
    vehicle_age_group = factor(vehicle_age_group,
      levels = c("Invalid", "0-5 years", "6-10 years", "11-20 years",
                 "21-30 years", "31-40 years", "41-50 years", "50+ years")),
    driver_age_group = case_when(
      driver_age <= 25 ~ "18-25 years",
      driver_age <= 35 ~ "26-35 years",
      driver_age <= 50 ~ "36-50 years",
      driver_age <= 65 ~ "51-65 years",
      TRUE ~ "65+ years"
    ),
    driver_age_group = factor(driver_age_group,
      levels = c("18-25 years", "26-35 years", "36-50 years", "51-65 years", "65+ years")),
    age_driving_licence_years = year - age_driving_licence,
    driving_experience_group = case_when(
      age_driving_licence_years < 0 ~ "Invalid",
      age_driving_licence_years <= 10 ~ "0-10 years",
      age_driving_licence_years <= 20 ~ "11-20 years",
      age_driving_licence_years <= 30 ~ "21-30 years",
      age_driving_licence_years <= 40 ~ "31-40 years",
      TRUE ~ "40+ years"
    ),
    driving_experience_group = factor(driving_experience_group,
      levels = c("Invalid", "0-10 years", "11-20 years", "21-30 years", "31-40 years", "40+ years"))
  ) %>%
  filter(!is.na(profitability), !is.na(age_driving_licence), !is.na(vehicle_age),
         !is.na(fuel_type), !is.na(vehicle_value))
```

**Portfolio Size:** `r format(nrow(insurance_clean), big.mark=",")` policies analyzed

**What We Examined:**

- Premium amounts and claim costs
- Vehicle characteristics (age, type, fuel)
- Customer location and policy type
- Profitability (Premium minus Claims)

::: notes
We analyzed over 600,000 active insurance policies. For each policy, we calculated profitability as the difference between premium collected and claims paid out. The dataset includes vehicle details, customer location, policy coverage type, and claims history. This large sample size gives us confidence in identifying meaningful patterns that can guide business decisions.
:::

## Portfolio Health Snapshot

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 5
#| message: false
pacman::p_load(patchwork)

# Calculate key metrics
pct_no_claims <- mean(insurance_clean$total_incurred == 0) * 100
pct_profitable <- mean(insurance_clean$profitability > 0) * 100
median_premium <- median(insurance_clean$total_premium)
median_profit <- median(insurance_clean$profitability)

# Premium plot
p1 <- ggplot(insurance_clean, aes(x = total_premium)) +
  geom_histogram(bins = 50, fill = "#2E86AB", alpha = 0.8) +
  geom_vline(xintercept = median_premium, linetype = "dashed",
             color = "#A23B72", linewidth = 1) +
  annotate("text", x = median_premium * 1.5, y = Inf,
           label = paste0("Typical: €", round(median_premium)),
           vjust = 2, hjust = 0, size = 4, color = "#A23B72") +
  scale_x_continuous(labels = comma, limits = c(0, 2000)) +
  labs(title = "Most Pay €300-€600 Annually",
       x = "Premium (€)", y = "Number of Policies") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(size = 11, face = "bold"))

# Claims plot
p2 <- ggplot(insurance_clean, aes(x = total_incurred)) +
  geom_histogram(bins = 50, fill = "#F18F01", alpha = 0.8) +
  annotate("text", x = 1000, y = Inf,
           label = paste0(round(pct_no_claims), "% Have No Claims"),
           vjust = 2, hjust = 0.5, size = 4, color = "#C73E1D") +
  scale_x_continuous(labels = comma, limits = c(0, 2000)) +
  labs(title = "9 in 10 Policies Have No Claims",
       x = "Claim Cost (€)", y = "Number of Policies") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(size = 11, face = "bold"))

# Profitability plot
p3 <- ggplot(insurance_clean, aes(x = profitability)) +
  geom_histogram(bins = 50, fill = "#06A77D", alpha = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 1) +
  geom_vline(xintercept = median_profit, linetype = "dashed",
             color = "#005F73", linewidth = 1) +
  annotate("text", x = median_profit * 1.5, y = Inf,
           label = paste0(round(pct_profitable), "% Profitable"),
           vjust = 2, hjust = 0, size = 4, color = "#005F73") +
  scale_x_continuous(labels = comma, limits = c(-500, 1500)) +
  labs(title = "Portfolio Is Profitable Overall",
       x = "Profit per Policy (€)", y = "Number of Policies") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(size = 11, face = "bold"))

p1 | p2 | p3
```

::: notes
Three key takeaways: (1) Most customers pay €300-€600 in annual premiums, with a small premium segment paying much more. (2) 88% of policies generate no claims - these customers fund the 12% who do claim. (3) The portfolio is healthy with most policies profitable, but there's significant variation. Our challenge is to identify what makes policies profitable and reduce the loss-making segment.
:::

## Pricing Model Needs Enhancement

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5.5
set.seed(123)
sample_data <- insurance_clean %>% sample_n(min(5000, nrow(insurance_clean)))

ggplot(sample_data, aes(x = total_premium, y = total_incurred)) +
  geom_point(alpha = 0.2, color = "#2E86AB", size = 1.5) +
  geom_smooth(method = "loess", color = "#C73E1D", se = TRUE, linewidth = 1.2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              color = "darkgreen", linewidth = 0.8) +
  annotate("text", x = 1200, y = 1200,
           label = "Perfect pricing\n(Premium = Claims)",
           color = "darkgreen", size = 3.5, hjust = 0) +
  annotate("text", x = 800, y = 200,
           label = "Weak connection between\npremium and actual claims",
           color = "#C73E1D", size = 4, hjust = 0.5) +
  scale_x_continuous(labels = comma, limits = c(0, 2000)) +
  scale_y_continuous(labels = comma, limits = c(0, 2000)) +
  labs(title = "Higher Premiums Don't Always Mean Higher Claims",
       subtitle = "Our pricing model is missing important risk factors",
       x = "Premium Charged (€)", y = "Actual Claim Cost (€)") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

::: notes
We charge higher premiums for policies we believe are riskier. However, the weak connection between premium and actual claims suggests we're missing important risk factors. Some high-premium customers rarely claim, while some low-premium customers have expensive claims. This represents both a risk (underpricing some segments) and an opportunity (we may be overcharging low-risk customers who could churn). Action: Enhance pricing model with additional factors like vehicle type, location, and driver history.
:::

## Fuel Type Shows Minimal Impact

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5
insurance_clean %>%
  group_by(fuel_type) %>%
  summarise(
    avg_profit = mean(profitability),
    n = n(),
    .groups = "drop"
  ) %>%
  filter(n > 100) %>%
  ggplot(aes(x = reorder(fuel_type, avg_profit), y = avg_profit, fill = fuel_type)) +
  geom_col(alpha = 0.8, width = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  geom_text(aes(label = paste0("€", round(avg_profit))),
            hjust = -0.2, size = 4) +
  coord_flip() +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0.05, 0.15))) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Fuel Type Has Little Effect on Profitability",
       subtitle = "Average profit per policy is similar across fuel types",
       x = NULL, y = "Average Profit per Policy (€)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold"))
```

::: notes
Diesel, gasoline, and hybrid vehicles show similar average profitability (within €50 of each other). Fuel type alone is not a strong predictor of policy performance. Implication: Current fuel-based pricing differences are appropriate; focus optimization efforts on factors that matter more, like policy type and customer tenure.
:::

## Location Matters: Area U Outperforms Area R

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5
area_summary <- insurance_clean %>%
  group_by(circulation_area) %>%
  summarise(
    mean_profit = mean(profitability),
    median_profit = median(profitability),
    n = n(),
    .groups = "drop"
  )

portfolio_avg <- mean(insurance_clean$profitability)

ggplot(area_summary, aes(x = reorder(circulation_area, mean_profit),
                         y = mean_profit, fill = circulation_area)) +
  geom_col(alpha = 0.85, width = 0.6) +
  geom_hline(yintercept = portfolio_avg, linetype = "dashed",
             color = "darkgray", linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "solid", color = "red", linewidth = 0.5) +
  geom_text(aes(label = paste0("€", round(mean_profit))),
            hjust = -0.2, size = 4.5, fontface = "bold") +
  annotate("text", x = 1.5, y = portfolio_avg,
           label = "Portfolio Average", vjust = -0.5,
           color = "darkgray", size = 3.5) +
  coord_flip() +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0.05, 0.15))) +
  scale_fill_manual(values = c("R" = "#F18F01", "U" = "#06A77D")) +
  labs(title = "Area U Generates Higher Profit Per Policy",
       subtitle = paste0("Difference: €", round(area_summary$mean_profit[2] - area_summary$mean_profit[1]), " per policy"),
       x = "Geographic Area", y = "Average Profit per Policy (€)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold"))
```

::: notes
Geographic location affects profitability. Area U generates higher average profit per policy than Area R. This difference likely reflects varying claim frequencies, repair costs, and driving conditions across regions. Recommendation: Implement location-based pricing adjustments and investigate root causes of Area R underperformance (higher claims? Lower premiums? Different customer mix?).
:::

:::

## Older Vehicles Carry Higher Loss Risk

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5.5
age_summary <- insurance_clean %>%
  filter(!is.na(vehicle_age_group), vehicle_age_group != "Invalid") %>%
  group_by(vehicle_age_group) %>%
  summarise(
    avg_profit = mean(profitability),
    pct_unprofitable = mean(profitability < 0) * 100,
    n = n(),
    .groups = "drop"
  )

ggplot(age_summary, aes(x = vehicle_age_group, y = avg_profit, fill = avg_profit)) +
  geom_col(alpha = 0.85) +
  geom_hline(yintercept = 0, linetype = "solid", color = "red", linewidth = 0.8) +
  geom_text(aes(label = paste0("€", round(avg_profit), "\n",
                               round(pct_unprofitable), "% loss")),
            vjust = ifelse(age_summary$avg_profit > 0, -0.3, 1.3),
            size = 3.5) +
  scale_fill_gradient2(low = "#C73E1D", mid = "#F4A261", high = "#06A77D",
                       midpoint = 250) +
  scale_y_continuous(labels = comma) +
  labs(title = "Vehicles Over 40 Years Show Higher Loss Rates",
       subtitle = "Average profit and % of unprofitable policies by vehicle age",
       x = "Vehicle Age", y = "Average Profit per Policy (€)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

::: notes
Older vehicles (40+ years) show lower average profitability and higher rates of unprofitable policies. While most old-vehicle policies are still profitable, the increased risk of extreme losses requires attention. Recommendation: Consider stricter underwriting criteria or adjusted premiums for very old vehicles to reflect their higher risk profile.
:::

## Comprehensive Coverage Drives Higher Profits

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5
policy_summary <- insurance_clean %>%
  group_by(policy_type) %>%
  summarise(
    avg_profit = mean(profitability),
    median_profit = median(profitability),
    n = n(),
    .groups = "drop"
  )

ggplot(policy_summary, aes(x = reorder(policy_type, avg_profit),
                           y = avg_profit, fill = policy_type)) +
  geom_col(alpha = 0.85, width = 0.6) +
  geom_hline(yintercept = 0, linetype = "solid", color = "red", linewidth = 0.8) +
  geom_text(aes(label = paste0("€", round(avg_profit))),
            hjust = -0.2, size = 5, fontface = "bold") +
  coord_flip() +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0.05, 0.2))) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Comprehensive Policies Are Most Profitable",
       subtitle = "Average profit per policy by coverage type",
       x = "Policy Type", y = "Average Profit per Policy (€)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold"))
```

::: notes
Comprehensive policies (COMP_N, COMP_E) generate significantly higher profit per policy than basic third-party (TP) coverage. This is our most actionable finding: comprehensive coverage provides better value for both customers and the company. Strategy: Incentivize customers to upgrade from third-party to comprehensive coverage through bundling, discounts, or highlighting the value proposition.
:::

## Customer Tenure Drives Profit

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5.5
#| warning: false
set.seed(123)
sample_data <- insurance_clean %>% sample_n(min(5000, nrow(insurance_clean)))

ggplot(sample_data, aes(x = total_exposure, y = profitability)) +
  geom_point(alpha = 0.2, color = "#2E86AB", size = 1.5) +
  geom_smooth(method = "loess", color = "#06A77D", se = TRUE, linewidth = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  annotate("text", x = 0.7, y = 800,
           label = "Longer-term customers\nare more profitable",
           color = "#06A77D", size = 4.5, hjust = 0, fontface = "bold") +
  scale_x_continuous(breaks = seq(0, 1, 0.2)) +
  scale_y_continuous(labels = comma) +
  labs(title = "Longer-Term Customers Generate Significantly More Profit",
       subtitle = "Each additional year of coverage adds to profitability",
       x = "Years of Coverage", y = "Profit per Policy (€)") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

::: notes
Customers who stay with us longer generate significantly more profit. This is the strongest relationship we found in the analysis. While this is partly because longer policies collect more premium, it highlights the critical importance of customer retention. Priority: Invest in retention programs, especially in the first 2 years when acquisition costs are highest. Reduce churn through loyalty programs, excellent service, and competitive renewal pricing.
:::

## Bonus Score Patterns

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5
bonus_summary <- insurance_clean %>%
  group_by(bonus_score) %>%
  summarise(
    avg_profit = mean(profitability),
    n = n(),
    .groups = "drop"
  )

ggplot(bonus_summary, aes(x = bonus_score, y = avg_profit, fill = bonus_score)) +
  geom_col(alpha = 0.85, width = 0.6) +
  geom_hline(yintercept = 0, linetype = "solid", color = "red", linewidth = 0.8) +
  geom_text(aes(label = paste0("€", round(avg_profit), "\nn=", format(n, big.mark=","))),
            vjust = -0.3, size = 4) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0.05, 0.15))) +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Claims History Shows Mixed Pattern",
       subtitle = "B (poor history) customers pay higher premiums that offset their claims",
       x = "Bonus Score (G=Good, N=Neutral, B=Bad)", y = "Average Profit per Policy (€)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold"))
```

::: notes
Customers with poor claims history (B) appear more profitable than those with good history (G). However, the B group is much smaller, and they pay higher premiums that more than offset their claims. This finding requires careful interpretation due to small sample size. Monitor this pattern as more data accumulates, but don't over-interpret given the limited B-group data.
:::

## Claims Volatility by Segment

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5
volatility_data <- insurance_clean %>%
  filter(total_incurred > 0) %>%
  group_by(fuel_type) %>%
  summarise(
    mean_claim = mean(total_incurred),
    sd_claim = sd(total_incurred),
    cv = sd_claim / mean_claim,
    n = n(),
    .groups = "drop"
  ) %>%
  filter(n > 1000)

ggplot(volatility_data, aes(x = reorder(fuel_type, cv), y = cv, fill = fuel_type)) +
  geom_col(alpha = 0.85, width = 0.6) +
  geom_text(aes(label = round(cv, 2)), hjust = -0.2, size = 4.5, fontface = "bold") +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Claim Costs Are Most Unpredictable for Gasoline Vehicles",
       subtitle = "Higher values = more unpredictable claims (harder to forecast reserves)",
       x = "Fuel Type", y = "Claim Cost Predictability Score") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold"))
```

::: notes
When gasoline vehicle owners make claims, the costs vary widely—making it harder to predict reserves needed. Diesel vehicles show more consistent claim costs, reducing our financial uncertainty. Implication: Set aside larger reserves for gasoline vehicle segment; consider reinsurance for extreme cases to protect against unexpected large claims.
:::

## Key Insights Summary

**What Drives Profitability:**

1.  **Customer Loyalty Matters Most** — Longer-term customers are significantly more profitable
2.  **Comprehensive Coverage Wins** — These policies outperform basic third-party coverage
3.  **Location Makes a Difference** — Area U generates higher profit than Area R
4.  **Vehicle Age Affects Risk** — Vehicles over 40 years show higher loss rates

**Where We Have Gaps:**

5.  **Pricing Needs Refinement** — Weak connection between premiums charged and actual claims suggests we're missing important risk factors

**Portfolio Health:**

- Overall portfolio is profitable with most customers paying €300-€600 annually
- 9 out of 10 policies have no claims
- Gasoline vehicles have less predictable claim costs than diesel

::: notes
Five key takeaways for business strategy: (1) Invest in customer retention programs - loyalty drives profit. (2) Promote comprehensive coverage upgrades. (3) Adjust pricing by location. (4) Review underwriting for very old vehicles. (5) Enhance pricing model with additional risk factors. The portfolio is fundamentally healthy, but these improvements can boost profitability and reduce risk.
:::

## Strategic Recommendations

**High Priority Actions:**

**1. Boost Customer Retention** - Launch loyalty programs targeting first 2-year customers - Implement competitive renewal pricing to reduce churn - *Impact: Longer tenure = higher profitability*

**2. Promote Comprehensive Coverage** - Create upgrade incentives for third-party customers - Bundle comprehensive with value-added services - *Impact: Comprehensive policies significantly more profitable*

**Medium Priority Actions:**

**3. Location-Based Pricing** - Adjust Area R premiums to reflect higher risk - Investigate root causes of regional differences - *Impact: Better risk-reward balance across regions*

**4. Vehicle Age Underwriting** - Tighten criteria for vehicles over 40 years old - Consider specialized pricing for vintage vehicles - *Impact: Reduce exposure to high-loss segments*

**5. Enhance Pricing Model** - Incorporate additional risk factors beyond current model - Improve premium-to-claims alignment - *Impact: Better risk prediction and pricing accuracy*

::: notes
Prioritized by impact and feasibility. High-priority actions (retention and comprehensive coverage promotion) offer the clearest path to improved profitability. Medium-priority actions address specific risk segments. All recommendations are grounded in the data analysis and represent actionable next steps for portfolio management.
:::
