---
title: "Take-home Exercise 1 (Presentation): Motor Insurance Portfolio Analysis"
subtitle: "Visual Analytics of Profitability and Risk"
author: "Bui Tien Thanh"
date: "May 9, 2026"
format:
  revealjs:
    theme: simple
    transition: slide
    slide-number: true
    chalkboard: true
---

## Contents {.smaller}

1.  Introduction & Objectives
2.  Data Overview
3.  Normality Assessment
4.  Premium Distribution
5.  Claims Distribution
6.  Profitability Patterns
7.  Premium vs Claims Relationship
8.  Profitability by Policy Type
9.  Total Exposure vs Profitability
10. Profitability by Bonus Score
11. Profitability by Vehicle Age
12. Geographic Analysis
13. Volatility Assessment
14. Statistical Validation
15. Key Findings
16. Recommendations

::: notes
This presentation analyzes a Spanish motor insurance portfolio using visual analytics techniques. We examine univariate distributions of key variables and bivariate relationships to identify factors influencing profitability and volatility. The analysis employs ggplot2 and statistical testing to validate findings and provide actionable insights for portfolio management.
:::

## Introduction & Objectives

**Dataset:** Spanish motor insurance portfolio

**Key Questions:**

- What are the distribution patterns of premiums and claims?
- Which factors are associated with portfolio profitability?
- What patterns are observed in claims volatility across segments?

**Methods:** Univariate and bivariate visual analytics using ggplot2 and ggstatsplot

::: notes
The study analyzes real-world insurance data to understand profitability drivers and risk patterns. We use tidyverse for data preparation and ggplot2 extensions for visualization. The objective is to identify actionable insights for pricing strategy and risk management through statistical validation of observed patterns.
:::

## Data Overview

```{r}
#| echo: false
#| message: false
pacman::p_load(tidyverse, ggplot2, readxl, ggstatsplot, scales, ggridges, ggdist, patchwork)

insurance_data <- read.csv(
  "../../data/Dataset of motor insurance portfolio.csv",
  sep = ";",
  dec = ".")

insurance_clean <- insurance_data %>%
  mutate(
    years_with_licence = year - age_driving_licence,
    profitability = total_premium - total_incurred,
    profit_ratio = ifelse(total_premium > 0, profitability / total_premium, NA),
    has_claim = ifelse(total_incurred > 0, "Yes", "No"),
    vehicle_age_group = case_when(
      vehicle_age < 0 ~ "Invalid",
      vehicle_age <= 5 ~ "0-5 years",
      vehicle_age <= 10 ~ "6-10 years",
      vehicle_age <= 20 ~ "11-20 years",
      vehicle_age <= 30 ~ "21-30 years",
      vehicle_age <= 40 ~ "31-40 years",
      vehicle_age <= 50 ~ "41-50 years",
      TRUE ~ "50+ years"
    ),
    vehicle_age_group = factor(vehicle_age_group,
      levels = c("Invalid", "0-5 years", "6-10 years", "11-20 years",
                 "21-30 years", "31-40 years", "41-50 years", "50+ years")),
    driver_age_group = case_when(
      driver_age <= 25 ~ "18-25 years",
      driver_age <= 35 ~ "26-35 years",
      driver_age <= 50 ~ "36-50 years",
      driver_age <= 65 ~ "51-65 years",
      TRUE ~ "65+ years"
    ),
    driver_age_group = factor(driver_age_group,
      levels = c("18-25 years", "26-35 years", "36-50 years", "51-65 years", "65+ years")),
    age_driving_licence_years = year - age_driving_licence,
    driving_experience_group = case_when(
      age_driving_licence_years < 0 ~ "Invalid",
      age_driving_licence_years <= 10 ~ "0-10 years",
      age_driving_licence_years <= 20 ~ "11-20 years",
      age_driving_licence_years <= 30 ~ "21-30 years",
      age_driving_licence_years <= 40 ~ "31-40 years",
      TRUE ~ "40+ years"
    ),
    driving_experience_group = factor(driving_experience_group,
      levels = c("Invalid", "0-10 years", "11-20 years", "21-30 years", "31-40 years", "40+ years"))
  ) %>%
  filter(!is.na(profitability), !is.na(age_driving_licence), !is.na(vehicle_age),
         !is.na(fuel_type), !is.na(vehicle_value))
```

**Dataset Size:** `r format(nrow(insurance_clean), big.mark=",")` policies (from `r format(nrow(insurance_data), big.mark=",")` original records)

**Key Variables:** Premium, Claims, Vehicle Type, Geographic Area, Vehicle Age

**Derived Metrics:** Profitability, Volatility (CV)

::: notes
The dataset contains over 600,000 insurance policies with information on premiums, claims, vehicle characteristics, and geographic location. We engineered profitability as the difference between premium and claim cost. Missing values were minimal and removed. This comprehensive dataset enables robust statistical analysis of factors affecting insurance portfolio performance.
:::

## Normality Assessment

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6
#| message: false
pacman::p_load(patchwork)

p_qq1 <- ggplot(insurance_clean, aes(sample = total_premium)) +
  stat_qq(alpha = 0.4) +
  stat_qq_line(color = "red") +
  labs(title = "Q-Q Plot: Premium", x = "Theoretical", y = "Sample") +
  theme_minimal(base_size = 12)

p_qq2 <- ggplot(insurance_clean, aes(sample = total_incurred)) +
  stat_qq(alpha = 0.4) +
  stat_qq_line(color = "red") +
  labs(title = "Q-Q Plot: Claims", x = "Theoretical", y = "Sample") +
  theme_minimal(base_size = 12)

p_qq3 <- ggplot(insurance_clean, aes(sample = profitability)) +
  stat_qq(alpha = 0.4) +
  stat_qq_line(color = "red") +
  labs(title = "Q-Q Plot: Profitability", x = "Theoretical", y = "Sample") +
  theme_minimal(base_size = 12)

(p_qq1 | p_qq2 | p_qq3)
```

::: notes
Q-Q plots show clear deviations from normality for all variables. Premium and claims are right-skewed; profitability shows left-tail deviation. Non-parametric tests (Kruskal-Wallis, Mann-Whitney) are appropriate for group comparisons.
:::

## Premium Distribution

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5
ggplot(insurance_clean, aes(x = total_premium)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 50, fill = "steelblue", alpha = 0.7) +
  geom_density(color = "red", linewidth = 1) +
  scale_x_continuous(labels = comma) +
  labs(
    title = "Distribution of Premium",
    x = "Premium (€)",
    y = "Density"
  ) +
  theme_minimal(base_size = 14)
```

::: notes
Premium distribution is strongly right-skewed with most policies concentrated between €300-€600. The long tail indicates a small proportion of high-premium policies. This pattern is typical for insurance portfolios where standard coverage dominates. The skewness suggests potential for segmentation strategies targeting different premium tiers with tailored products and pricing approaches.
:::

## Claims Distribution

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5
ggplot(insurance_clean, aes(x = total_incurred)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, fill = "coral", alpha = 0.7) +
  geom_density(color = "darkred", linewidth = 1) +
  scale_x_continuous(labels = comma) +
  labs(title = "Distribution of Claim Costs", x = "Claim Cost (€)", y = "Density") +
  theme_minimal(base_size = 14)
```

::: notes
Claims distribution shows extreme right-skewness with high concentration at zero, indicating most policies have no claims. Among policies with claims, costs vary widely with a long tail of expensive claims. This pattern reflects the fundamental insurance principle where few policyholders generate claims. The high zero-inflation suggests opportunities for no-claims bonus programs to retain low-risk customers.
:::

## Profitability Patterns

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5
ggplot(insurance_clean, aes(x = profitability)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 50, fill = "forestgreen", alpha = 0.7) +
  geom_density(color = "darkgreen", linewidth = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  scale_x_continuous(labels = comma) +
  labs(
    title = "Distribution of Profitability",
    subtitle = "Profitability = Premium - Claims",
    x = "Profitability (€)",
    y = "Density"
  ) +
  theme_minimal(base_size = 14)
```

::: notes
Overall profitability distribution centers around positive values, indicating a healthy portfolio. However, substantial variance exists with a significant proportion of unprofitable policies. The distribution is slightly right-skewed, suggesting more extreme profits than losses. This pattern validates the portfolio's viability while highlighting the need to identify and mitigate loss-making segments through improved risk assessment and pricing adjustments.
:::

## Premium vs Claims Relationship

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5
cor_value <- cor(insurance_clean$total_premium, insurance_clean$total_incurred, method = "spearman")

ggplot(insurance_clean %>% sample_n(min(5000, nrow(insurance_clean))),
       aes(x = total_premium, y = total_incurred)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(title = "Premium vs Claim Cost Relationship",
       subtitle = paste0("Spearman correlation: ", round(cor_value, 3)),
       x = "Premium (€)", y = "Claim Cost (€)") +
  theme_minimal(base_size = 14)
```

::: notes
The scatter plot reveals a weak positive correlation between premiums and claim costs (Spearman ρ ≈ 0.1-0.2). This suggests current pricing may not fully capture underlying risk factors. A stronger correlation would indicate better risk-based pricing. The weak relationship highlights opportunities to refine pricing models by incorporating additional risk factors such as vehicle type, driver characteristics, and geographic location to better align premiums with expected claims.
:::

## Profitability by Fuel Type (Detailed)

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5
insurance_clean %>%
  filter(profitability >= -2000, profitability <= 2000) %>%
  ggplot(aes(x = fuel_type, y = profitability, fill = fuel_type)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.25) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Profitability by Fuel Type",
    subtitle = "Filtered to profitability between -€2,000 and €2,000 for readability",
    x = "Fuel Type",
    y = "Profitability (€)"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

::: notes
Boxplot shows the distribution of profitability across fuel types with extreme outliers filtered for clarity. The median profitability is similar across fuel types, with slight variations. The dashed red line at zero helps identify unprofitable policies.
:::

## Geographic Analysis

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5
insurance_clean %>%
  group_by(circulation_area) %>%
  summarise(mean_profit = mean(profitability), n = n()) %>%
  ggplot(aes(x = reorder(circulation_area, mean_profit), y = mean_profit, fill = circulation_area)) +
  geom_col() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(title = "Average Profitability by Circulation Area", x = "Area", y = "Mean Profitability (€)") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

::: notes
Geographic location is significantly associated with profitability, with substantial variation across areas. Some regions consistently generate higher profits while others underperform. Statistical tests confirm these differences are significant (p \< 0.001). Regional variations likely reflect differences in claim frequency, repair costs, fraud rates, and driving conditions. This finding supports geographic risk-based pricing and targeted market strategies for high-performing regions.
:::

## Profitability by Region (Statistical Test)

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6
#| warning: false
insurance_clean %>%
  filter(profitability >= -2000, profitability <= 2000) %>%
  ggbetweenstats(
    data = .,
    x = circulation_area,
    y = profitability,
    type = "nonparametric",
    pairwise.comparisons = TRUE,
    p.adjust.method = "bonferroni",
    title = "Profitability Distribution by Circulation Area"
  )
```

::: notes
Statistical validation shows area U has slightly higher median profitability than area R. The difference is statistically significant but the effect size is very small, indicating circulation area alone doesn't strongly explain profitability differences.
:::

## Raincloud Plot: Profitability by Area

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6
#| warning: false
# Data for plotting only
plot_data <- insurance_clean %>%
  filter(profitability >= -2000, profitability <= 2000)

# Sample points for cleaner display
set.seed(123)
point_data <- plot_data %>%
  group_by(circulation_area) %>%
  slice_sample(n = 1500) %>%
  ungroup()

# Median values
median_data <- plot_data %>%
  group_by(circulation_area) %>%
  summarise(
    median_profit = median(profitability, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(plot_data, aes(x = circulation_area, y = profitability,
                      fill = circulation_area, color = circulation_area)) +
  stat_halfeye(
    adjust = 0.7,
    width = 0.6,
    justification = -0.25,
    .width = 0,
    point_colour = NA,
    alpha = 0.45
  ) +
  geom_boxplot(
    width = 0.12,
    outlier.shape = NA,
    alpha = 0.5,
    color = "black"
  ) +
  geom_jitter(
    data = point_data,
    width = 0.08,
    alpha = 0.10,
    size = 0.6,
    show.legend = FALSE
  ) +
  geom_point(
    data = median_data,
    aes(x = circulation_area, y = median_profit),
    inherit.aes = FALSE,
    color = "black",
    size = 2.5
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Raincloud Plot: Profitability by Circulation Area",
    subtitle = "Filtered to profitability between -€2,000 and €2,000 for readability",
    x = "Circulation Area",
    y = "Profitability (€)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank()
  )
```

::: notes
Raincloud plots combine density curves, boxplots, and individual data points to show complete distributional information. This advanced visualization reveals not just central tendency but also spread, skewness, and outliers across geographic areas, enabling more nuanced risk assessment.
:::

## Profitability by Vehicle Age

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6
insurance_clean %>%
  filter(
    profitability >= -2000,
    profitability <= 2000,
    !is.na(vehicle_age_group)
  ) %>%
  ggplot(aes(
    x = vehicle_age_group,
    y = profitability,
    fill = vehicle_age_group
  )) +
  geom_violin(alpha = 0.7, trim = TRUE) +
  geom_boxplot(width = 0.2, alpha = 0.5, outlier.alpha = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Profitability by Vehicle Age Group",
    subtitle = "Filtered to profitability between -€2,000 and €2,000 for readability",
    x = "Vehicle Age Group",
    y = "Profitability (€)"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

::: notes
Profitability varies across vehicle age groups. Older vehicles, especially 41-50 and 50+ groups, show wider negative tails, suggesting higher downside risk. Most policies across all age groups have profitability close to zero or moderately positive.
:::

## Profitability by Policy Type

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6
#| warning: false
ggbetweenstats(
  data = insurance_clean,
  x = policy_type,
  y = profitability,
  type = "nonparametric",
  pairwise.comparisons = TRUE,
  p.adjust.method = "bonferroni",
  title = "Profitability Distribution by Policy Type"
)
```

::: notes
Policy type shows the most meaningful group difference in the analysis. COMP_N and COMP_E (comprehensive policies) have the highest median profitability, while TP (basic third-party) has the lowest. This is likely because comprehensive policies generate higher premiums that offset claim costs. Although the effect size is still relatively small, this is the most actionable finding for pricing strategy.
:::

## Total Exposure vs Profitability

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5
#| warning: false
set.seed(123)
cor_exp <- cor(insurance_clean$total_exposure, insurance_clean$profitability,
               method = "spearman", use = "complete.obs")

insurance_clean %>%
  sample_n(min(5000, nrow(insurance_clean))) %>%
  ggplot(aes(x = total_exposure, y = profitability)) +
  geom_point(alpha = 0.25, color = "orange") +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Total Exposure vs Profitability",
    subtitle = paste0("Spearman correlation: ", round(cor_exp, 3)),
    x = "Total Exposure (years)",
    y = "Profitability (€)"
  ) +
  theme_minimal(base_size = 14)
```

::: notes
Total exposure has the strongest correlation with profitability (Spearman ρ = 0.565) — the highest of any variable in the analysis. Longer-active policies earn more premium, which directly increases profitability. However, this relationship is partly structural: profitability is constructed from premium, which scales with exposure. Future analysis should use exposure-adjusted metrics for fairer comparisons.
:::

## Profitability by Bonus Score

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6
#| warning: false
ggbetweenstats(
  data = insurance_clean,
  x = bonus_score,
  y = profitability,
  type = "nonparametric",
  pairwise.comparisons = TRUE,
  p.adjust.method = "bonferroni",
  title = "Profitability Distribution by Bonus Score"
)
```

::: notes
Bonus score shows a counterintuitive pattern: the B group (poor claims history) has the highest median profitability (\~€333), while G (good claims history) has the lowest (\~€229). This may reflect that B-group policyholders pay higher premiums that more than offset their claims. However, the B and N groups are much smaller than G, so results should be interpreted cautiously. The effect size is small.
:::

## Volatility Assessment

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5
insurance_clean %>%
  group_by(fuel_type) %>%
  summarise(mean_claim = mean(total_incurred), sd_claim = sd(total_incurred), cv = sd_claim / mean_claim, n = n()) %>%
  filter(n > 1000) %>%
  ggplot(aes(x = reorder(fuel_type, cv), y = cv, fill = fuel_type)) +
  geom_col() +
  coord_flip() +
  labs(title = "Claims Volatility by Fuel Type", subtitle = "Coefficient of Variation", x = "Fuel Type", y = "CV") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

::: notes
Claims volatility, measured by coefficient of variation, differs substantially across fuel types. Higher volatility indicates greater uncertainty in claim costs, representing increased risk exposure. Certain segments show particularly high volatility, requiring larger reserves and potentially higher risk premiums. Understanding volatility patterns enables better capital allocation and risk management strategies to ensure portfolio stability and solvency.
:::

## Statistical Validation

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5
#| warning: false
insurance_clean %>%
  ggbetweenstats(
    x = fuel_type,
    y = profitability,
    type = "nonparametric",
    pairwise.comparisons = FALSE,
    title = "Profitability Differences: Statistical Test"
  )
```

::: notes
Non-parametric Kruskal-Wallis tests confirm that profitability differences across vehicle types are statistically significant (p \< 0.001). The effect size indicates practical significance beyond statistical significance. Pairwise comparisons reveal which specific vehicle types differ significantly from each other. This rigorous statistical validation ensures our findings are not due to random chance and provides confidence for implementing data-driven pricing and underwriting decisions.
:::

## Key Findings {.smaller}

**Univariate:**

- Premium right-skewed (most policies \< €500); claims 88.1% zero-inflated; profitability has long negative left tail
- Portfolio dominated by diesel (63.9%), CC policy type (57.9%), good claims history (bonus G)

**Bivariate:**

- Total exposure strongest predictor of profitability (Spearman ρ = 0.565) — longer policies earn more premium
- Policy type most meaningful group difference: COMP_N/COMP_E more profitable than TP
- All other categorical factors (fuel, area, vehicle age, driver age, experience, municipality, bonus score) are statistically significant but with very small effect sizes
- Premium vs claims: weak relationship (ρ = 0.217) — premium alone does not predict claim cost

**Volatility:**

- Gasoline vehicles and Area R show highest claim cost volatility (CV)
- All CV measures inflated by zero-inflation; treat as relative risk indicators

::: notes
Key takeaway: no single factor strongly explains profitability. Policy type and total exposure are the most actionable variables. Statistical significance across many variables is largely driven by the large sample size rather than strong practical effects.
:::

## Recommendations {.smaller}

**1. Policy Type Pricing**

- Comprehensive policies (COMP_N, COMP_E) are more profitable — review whether TP premiums adequately cover liability exposure

**2. Claim Severity Management**

- A small number of extreme claims drive most losses — consider differentiated coverage terms or reinsurance for high-severity segments

**3. Vehicle Age Risk**

- Older vehicles (41–50 years, 50+) show wider negative profitability tails — tighten underwriting criteria for very old vehicles

**4. Geographic Volatility**

- Area R has higher claim volatility than Area U — adjust pricing and reserving assumptions by region

**5. Exposure-Adjusted Analysis**

- Total exposure strongly correlates with profitability by construction — use exposure-adjusted metrics for fairer segment comparisons in future analysis

::: notes
Recommendations are grounded in the actual effect sizes observed. Policy type and claim severity management are the highest-priority actions. Geographic and vehicle age adjustments are secondary but statistically supported.
:::
